---
title: "Find best 1-class model before estimating growth mixture models on ubelix"
author: "Lorenz Leuenberger"
date: "`r date()`"
title-block-banner: "#e4003c"
format: 
  html:
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
editor_options: 
  chunk_output_type: inline
---

# Introduction

I would like to first test whether we can use polynomials and splines with GMM.
Then I would like to find the best pattern for the 1-class model.

## Import data

```{r}


library(here)
library(lcmm) # for growth mixture models
library(splines) # for splines
library(tidyverse)
library(parallel)
library(mgcv) # for penalized splines

data <- read.csv(here("data", "processed", "Leicester_anthro_pseudonymized_18.csv"))
interpolated_ukwho_bmi <- read.csv(here("data", "interpolated-growth-references", "interpolated_ukwho_bmi.csv"))

```



# Test polynomials with GMM

1 latent class

```{r}

m_poly_1 <- hlme(fixed = bmi ~ poly(age_years, degree = 3),
                  random = ~ -1,
                  subject = "participant_id",
                  ng = 1,
                  data = data)

summary(m_poly_1)


data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_poly_1, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred",
         "lower_class1" = "pred.lower.Ypred",
         "upper_class1" = "pred.upper.Ypred")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_poly_1 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)

plot_m_poly_1

```


2 latent classes

```{r}


m_poly_2 <- hlme(fixed = bmi ~ poly(age_years, degree = 3),
               mixture = ~ poly(age_years, degree = 3),
               random = ~ -1,
               subject = "participant_id",
               ng = 2,
               B = m_poly_1,
               data = data)


data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_poly_2, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred_class1", "mean_class2" = "pred.Ypred_class2",
         "lower_class1" = "pred.lower.Ypred_class1", "lower_class2" = "pred.lower.Ypred_class2",
         "upper_class1" = "pred.upper.Ypred_class1", "upper_class2" = "pred.upper.Ypred_class2")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_poly_2 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)

plot_m_poly_2

```


Compare models

```{r}

summarytable(m_poly_1, m_poly_2)

```

# Test splines with GMM


## B splines

1 latent class

```{r}

m_bs_1 <- hlme(fixed = bmi ~  bs(age_years, knots = c(0.5875055, 7.251932, 14.35975), degree = 2),
                  random = ~ 1 + age_years,
                  subject = "participant_id",
                  ng = 1,
                  data = data)

summary(m_bs_1)

knots <- attr(splines::bs(data$age_years, degree = 2, df = 6), "knots")

summarytable(m_bs_1)

# plot
data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_bs_1, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred",
         "lower_class1" = "pred.lower.Ypred",
         "upper_class1" = "pred.upper.Ypred")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_bs_1 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)

plot_m_bs_1


```

2 latent classes

```{r}


m_bs_2 <- hlme(fixed = bmi ~ bs(age_years, knots = c(0.5875055, 7.251932, 14.35975), degree = 2),
               mixture = ~ bs(age_years, knots = c(0.5875055, 7.251932, 14.35975), degree = 2),
               random = ~ 1 + age_years,
               subject = "participant_id",
               ng = 3,
               B = m_bs_1,
               data = data)


data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_bs_2, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred_class1", "mean_class2" = "pred.Ypred_class2",
         "lower_class1" = "pred.lower.Ypred_class1", "lower_class2" = "pred.lower.Ypred_class2",
         "upper_class1" = "pred.upper.Ypred_class1", "upper_class2" = "pred.upper.Ypred_class2")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_bs_2 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)

plot_m_bs_2

summarytable(m_bs_2)

summary(m_bs_2)

```


Compare models

```{r}

summarytable(m_bs_1, m_bs_2)

```

## Natural cubic splines
1 latent class

```{r}

m_ns_1 <- hlme(fixed = bmi ~ splines::ns(age_years, df = 4),
                  random = ~ 1,
                  subject = "participant_id",
                  ng = 1,
                  data = data)

summary(m_ns_1)
summarytable(m_ns_1)

knots <- attr(splines::ns(data$age_years, df = 4), "knots")

data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_ns_1, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred",
         "lower_class1" = "pred.lower.Ypred",
         "upper_class1" = "pred.upper.Ypred")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_ns_1 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)

plot_m_ns_1


```


2 latent classes

```{r}


m_ns_2 <- hlme(fixed = bmi ~ splines::ns(age_years, df = 4),
               mixture = ~ splines::ns(age_years, df = 4),
               random = ~ 1,
               subject = "participant_id",
               ng = 2,
               B = m_ns_1,
               data = data)


data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

data_predict <- predictY(m_ns_2, data_predict, var.time = "age_years", draws = TRUE)

data_predict <- data.frame(data_predict[2], data_predict[1])

data_predict <- data_predict %>% 
  rename("mean_class1" = "pred.Ypred_class1", "mean_class2" = "pred.Ypred_class2",
         "lower_class1" = "pred.lower.Ypred_class1", "lower_class2" = "pred.lower.Ypred_class2",
         "upper_class1" = "pred.upper.Ypred_class1", "upper_class2" = "pred.upper.Ypred_class2")

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

plot_m_ns_2 <- data_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = class)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)

plot_m_ns_2

```


Compare models

```{r}

summarytable(m_ns_1, m_ns_2)

```


# Best 1-class lm model

## Estimate models



```{r}

m1a <- lm(bmi ~ age_years, data = data)

m1b <- lm(bmi ~ poly(age_years, degree = 2), data = data)

m1c <- lm(bmi ~ poly(age_years, degree = 3), data = data)

m1d <- lm(bmi ~ splines::bs(age_years, degree = 2, df = 4), data = data)

m1e <- lm(bmi ~ splines::bs(age_years, degree = 2, df = 5), data = data)

m1f <- lm(bmi ~ splines::bs(age_years, degree = 3, df = 5), data = data)

m1g <- lm(bmi ~ splines::bs(age_years, degree = 3, df = 6), data = data)

m1h <- lm(bmi ~ splines::ns(age_years, df = 3), data = data)

m1i <- lm(bmi ~ splines::ns(age_years, df = 4), data = data)

```

### Search for knot placement based on BIC and AIC

Define functions to perform gridsearch

```{r}

# Function to fit B-spline or Natural Spline model and return AIC/BIC
fit_spline_model <- function(knots, data, degree, spline_type = "bs") {
  if (spline_type == "bs") {
    model <- lm(bmi ~ bs(age_years, knots = knots, degree = degree), data = data)
  } else if (spline_type == "ns") {
    model <- lm(bmi ~ ns(age_years, knots = knots), data = data)
  } else {
    stop("Invalid spline_type. Use either 'bs' for B-splines or 'ns' for Natural splines.")
  }
  
  predicted <- predict(model, newdata = data)
  rmse <- sqrt(mean((data$bmi - predicted)^2))
  
  return(c(AIC = AIC(model), BIC = BIC(model), RMSE = rmse))
}

# Function to perform grid search with iterative refinement, supporting both B-splines and Natural splines
spline_gridsearch <- function(range_1, range_2, range_3 = NULL, num_knots, degree, data, iterations = 3, spline_type = "bs") {
  
  current_range_1 <- range_1
  current_range_2 <- range_2
  current_range_3 <- range_3
  
  best_knots <- NULL
  
  for (i in 1:iterations) {
    knots_1 <- seq(current_range_1[1], current_range_1[2], length.out = num_knots)
    knots_2 <- seq(current_range_2[1], current_range_2[2], length.out = num_knots)
    
    if (!is.null(current_range_3)) {
      knots_3 <- seq(current_range_3[1], current_range_3[2], length.out = num_knots)
      knots_df <- expand.grid(knot_1 = knots_1, knot_2 = knots_2, knot_3 = knots_3)
    } else {
      knots_df <- expand.grid(knot_1 = knots_1, knot_2 = knots_2)
    }
    
    no_cores <- detectCores() - 1
    cl <- makeCluster(no_cores)
    clusterExport(cl, list("fit_spline_model", "data"))
    clusterEvalQ(cl, library(splines))
    
    aic_bic_rmse_values <- parApply(cl, knots_df, 1, function(row) {
      if (!is.null(current_range_3)) {
        knots <- c(row[1], row[2], row[3])
      } else {
        knots <- c(row[1], row[2])
      }
      fit_spline_model(knots, data, degree, spline_type)
    })
    
    stopCluster(cl)
    
    results <- t(aic_bic_rmse_values)
    results <- as.data.frame(results)
    colnames(results) <- c("AIC", "BIC", "RMSE")
    results$knot_1 <- knots_df$knot_1
    results$knot_2 <- knots_df$knot_2
    if (!is.null(current_range_3)) {
      results$knot_3 <- knots_df$knot_3
    }
    
    best_knots_aic <- c(results$knot_1[which.min(results$AIC)], results$knot_2[which.min(results$AIC)], ifelse(!is.null(current_range_3), results$knot_3[which.min(results$AIC)], NA))
    best_knots_bic <- c(results$knot_1[which.min(results$BIC)], results$knot_2[which.min(results$BIC)], ifelse(!is.null(current_range_3), results$knot_3[which.min(results$BIC)], NA))
    best_knots_rmse <- c(results$knot_1[which.min(results$RMSE)], results$knot_2[which.min(results$RMSE)], ifelse(!is.null(current_range_3), results$knot_3[which.min(results$RMSE)], NA))
    
    best_knots <- best_knots_bic
    
    current_range_1 <- c(best_knots[1] - (knots_1[2] - knots_1[1]),
                         best_knots[1] + (knots_1[2] - knots_1[1]))
    current_range_2 <- c(best_knots[2] - (knots_2[2] - knots_2[1]),
                         best_knots[2] + (knots_2[2] - knots_2[1]))
    
    if (!is.null(current_range_3)) {
      current_range_3 <- c(best_knots[3] - (knots_3[2] - knots_3[1]),
                           best_knots[3] + (knots_3[2] - knots_3[1]))
    }
    
    cat("Iteration:", i, "\n")
    cat("Best BIC:", min(results$BIC), "; Best Knots:", best_knots, "\n")
  }
  
  return(best_knots)
}


```

Phases of BMI growth in children:

- infancy (0-1.5 years)
- childhood (1.5-12 years)
- puberty (12-16 years)

Based on previous knowledge of BMI growth and on the publication of [Nummi et al.](https://dx.doi.org/10.1080/02664763.2013.872232.) we search for the 2 knots between 0.5-2 years and 8-13 years. A possible 3rd knot is added between 14-17.5 years


```{r}

# quadratic b-splines

m1j_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), num_knots = 20, degree = 2, data = data, spline_type = "bs")

m1j <- lm(bmi ~ bs(age_years, knots = m1j_knots, degree = 2), data = data)

m1k_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), range_3 = c(14,17.5), num_knots = 20, degree = 2, data = data, spline_type = "bs")

m1k <- lm(bmi ~ bs(age_years, knots = m1k_knots, degree = 2), data = data)

# cubic b-splines

m1l_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), num_knots = 20, degree = 3, data = data, spline_type = "bs")

m1l <- lm(bmi ~ bs(age_years, knots = m1l_knots, degree = 3), data = data)

m1m_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), range_3 = c(14,17.5), num_knots = 20, degree = 3, data = data, spline_type = "bs")

m1m <- lm(bmi ~ bs(age_years, knots = m1m_knots, degree = 3), data = data)

# cubic natural splines

m1n_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), num_knots = 20, degree = 3, data = data, spline_type = "ns")

m1n <- lm(bmi ~ ns(age_years, knots = m1n_knots), data = data)

m1o_knots <- spline_gridsearch(range_1 = c(0.3,2), range_2 = c(7,13), range_3 = c(14,17.5), num_knots = 20, degree = 3, data = data, spline_type = "ns")

m1o <- lm(bmi ~ ns(age_years, knots = m1o_knots), data = data)

```


## Choose best model

Look at model summary

```{r}

compare_models <- function(models) {
  model_comparison <- data.frame(
    Model = character(),
    AIC = numeric(),
    BIC = numeric(),
    LogLik = numeric(),
    npm = numeric(),
    stringsAsFactors = FALSE
  )
  
  for (model_name in names(models)) {
    model <- models[[model_name]]
    
    model_comparison <- rbind(model_comparison, data.frame(
      Model = model_name,
      AIC = AIC(model),
      BIC = BIC(model),
      LogLik = as.numeric(logLik(model)),
      npm = length(coefficients(model)),
      stringsAsFactors = FALSE
    ))
  }
  
  model_comparison <- model_comparison %>% 
    arrange(BIC)
  
  return(model_comparison)
}

models <- list(
  m1a = m1a, m1b = m1b, m1c = m1c, m1d = m1d, m1e = m1e,
  m1f = m1f, m1g = m1g, m1h = m1h, m1i = m1i, m1j = m1j, m1k = m1k, m1l = m1l, m1m = m1m, m1n = m1n, m1o = m1o
)

model_comparison <- compare_models(models)

print(model_comparison)

```

Based on BIC:

- The best 1-class model is m1j with quadratic splines and 2 manually searched knots: 0.6650824 9.80449
- The second best 1-class model is m1k with quadratic splines and 2 manually searched knots: 0.6650824 10.34597 13.7964 (better AIC and lower LogLik)

## Plot models

```{r}

summary(m1j)

# plot
data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

m1j_predict <- predict(m1j, data_predict, se = TRUE)

m1j_predict <- data.frame(m1j_predict[2], m1j_predict[1])

m1j_predict <- m1j_predict %>% 
  mutate(lower_m1j = fit - 1.96* se.fit,
         upper_m1j = fit - 1.96* se.fit) %>% 
  rename("mean_m1j" = "fit")

m1j_predict <- m1j_predict %>% 
  cbind(data_predict) %>% 
  select(mean_m1j, lower_m1j, upper_m1j, age_years)

m1j_predict <- m1j_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

knots <- m1j_knots

plot_m1j <- m1j_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)+
  scale_x_continuous(breaks = c(0 , 2, 4, 6, 8, 10, 12, 14, 16, 18))+
  scale_y_continuous(limits = c(12, 25))+
  theme_minimal()

plot_m1j

```

The second best model was m1k: quadratic splines with 3 manually searched knot placements.

```{r}

summary(m1k)

# plot
data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

m1k_predict <- predict(m1k, data_predict, se = TRUE)

m1k_predict <- data.frame(m1k_predict[2], m1k_predict[1])

m1k_predict <- m1k_predict %>% 
  mutate(lower_m1k = fit - 1.96* se.fit,
         upper_m1k = fit - 1.96* se.fit) %>% 
  rename("mean_m1k" = "fit")

m1k_predict <- m1k_predict %>% 
  cbind(data_predict) %>% 
  select(mean_m1k, lower_m1k, upper_m1k, age_years)

m1k_predict <- m1k_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

knots <- attr(splines::bs(data$age_years, degree = 3, df = 6), "knots")

plot_m1k <- m1k_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)+
  scale_x_continuous(breaks = c(0 , 2, 4, 6, 8, 10, 12, 14, 16, 18))+
  scale_y_continuous(limits = c(12, 25))+
  theme_minimal()

plot_m1k

```

### Compare best models to expected pattern

```{r}

# penalized splines

mgam <- gam(bmi ~ s(age_years, bs = "bs", m = c(3,2)), data = data)

summary(mgam)

mgam_predict <- predict(mgam, data_predict, se = TRUE)

mgam_predict <- data.frame(mgam_predict[2], mgam_predict[1])

mgam_predict <- mgam_predict %>% 
  mutate(lower_gam = fit - 1.96 * se.fit,
         upper_gam = fit + 1.96 * se.fit) %>% 
  rename("mean_gam" = "fit")

mgam_predict <- mgam_predict %>% 
  cbind(data_predict) %>% 
  select(mean_gam, lower_gam, upper_gam, age_years)

mgam_predict <- mgam_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

# loess

mloess <- loess(bmi ~ age_years, data = data)

summary(mloess)

mloess_predict <- predict(mloess, data_predict, se = TRUE)

mloess_predict <- data.frame(mloess_predict[2], mloess_predict[1])

mloess_predict <- mloess_predict %>% 
  mutate(lower_loess = fit - 1.96* se.fit,
         upper_loess = fit - 1.96* se.fit) %>% 
  rename("mean_loess" = "fit")

mloess_predict <- mloess_predict %>% 
  cbind(data_predict) %>% 
  select(mean_loess, lower_loess, upper_loess, age_years)

mloess_predict <- mloess_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

# WHO growth references

who_predict <- interpolated_ukwho_bmi %>% 
  mutate(mean = M,
         age_years = age_days/365.25,
         model = case_when(sex == "m" ~ "who_boys",
                           sex == "f" ~ "who_girls"),
         lower = NA,
         upper = NA) %>% 
  select(age_years, model, mean, lower, upper)

models_predict <- rbind(m1k_predict, m1j_predict, mloess_predict, mgam_predict, who_predict) %>% 
  mutate(model = case_when(model == "m1k" ~ "quadratic, 3 manual knots",
                           model == "m1j" ~ "quadratic, 2 manual knots",
                           TRUE ~ model))


plot_models <- models_predict %>% 
  filter(model == "who_girls" | model == "who_boys") %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line() + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18)) +  
  scale_y_continuous(limits = c(12, 25)) +
  scale_color_manual(values = c("who_girls" = "red", "who_boys" = "blue"))+
  theme_minimal()+
  labs(y = "Mean BMI (kg/m^2)",
         x = "Age (years)")

plot_models

plot_models <- models_predict %>% 
  filter(model == "who_girls" | model == "who_boys" | model == "gam" | model == "loess") %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line() + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18)) +  
  scale_y_continuous(limits = c(12, 25)) +
  scale_color_manual(values = c("who_girls" = "red", "who_boys" = "blue", "gam" = "black", "loess" = "grey"))+
  theme_minimal()+
  labs(y = "Mean BMI (kg/m^2)",
         x = "Age (years)")

plot_models

plot_models <- models_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line() + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18)) +  
  scale_y_continuous(limits = c(12, 25)) +
  scale_color_manual(values = c("who_girls" = "red", "who_boys" = "blue", "gam" = "black", "loess" = "grey", "quadratic, 3 manual knots" = "green", "quadratic, 2 manual knots" = "yellow"))+
  theme_minimal()+
  labs(y = "Mean BMI (kg/m^2)",
         x = "Age (years)")

plot_models

```

# Best 1-class hlme model
## Estimate models

```{r}
sdf
# quadratic splines with 2 knots

m1r <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
            random = ~ -1,
            subject = "participant_id",
            ng = 1,
            nwg = FALSE,
            idiag = FALSE,
            data = data)

m1s <- hlme(fixed = bmi ~ splines::bs(age_years, degree = 2 , df = 4),
            random = ~ -1,
            subject = "participant_id",
            ng = 1,
            nwg = FALSE,
            idiag = FALSE,
            data = data)


# quadratic splines with 3 knots

m1t <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
            random = ~ -1,
            subject = "participant_id",
            ng = 1,
            nwg = FALSE,
            idiag = FALSE,
            data = data)

m1u <- hlme(fixed = bmi ~ splines::bs(age_years, degree = 2, df = 5),
            random = ~ -1,
            subject = "participant_id",
            ng = 1,
            nwg = FALSE,
            idiag = FALSE,
            data = data)

```


## Choose best model

Look at model summary

```{r}

summarytable(m1r, m1s, m1t, m1u, which = c("AIC", "BIC", "loglik", "npm"))

```

Choose best model

```{r}

select_best_model_of_class <- function(models, BIC = 0, p_Loglik = 0.05){
  
  invisible(capture.output(suppressMessages(suppressWarnings(models <- models[order(sapply(models, function(model) summarytable(model, which="npm")[1,1]))]))))
  
  best_model <- models[[1]]
  best_model_index <- 1
  
  for (i in 2:length(models)){
  
  cat("compare", names(models)[i], "to" , names(models)[best_model_index], "\n\n")
  
  new_model = models[[i]]
  
  summary_models <- summarytable(best_model, new_model, which = c("BIC", "loglik", "npm"))
  BIC_best_model <- summary_models[1,1]
  Loglik_best_model <- summary_models[1,2]
  number_parameters_best_model <- summary_models[1,3]
  BIC_new_model <- summary_models[2,1]
  Loglik_new_model <- summary_models[2,2]
  number_parameters_new_model <- summary_models[2,3]
  
  BIC_difference <- BIC_new_model - BIC_best_model
  Loglik_test <- pchisq(-2*(Loglik_best_model - Loglik_new_model),
                        df = abs(number_parameters_new_model-number_parameters_best_model),
                        lower.tail = FALSE)
  
  cat("\nBIC difference:\t\t\t", BIC_difference, "\n")
  cat("p-value Likelihood Ratio test:\t", Loglik_test, "\n\n")
  
  if (BIC_difference <= BIC & Loglik_test <= p_Loglik){
    
    cat(names(models)[i],"better than", names(models)[best_model_index], "\n\n")
    cat(paste(rep("-",80), collapse= ""), "\n\n")
    
    best_model <- new_model
    best_model_index <- i
  } else {
    
    cat(names(models)[i],"not better than", names(models)[best_model_index], "\n\n")
    cat(paste(rep("-",80), collapse= ""), "\n\n")
    
  }
  
  }

  cat("best model:", names(models)[best_model_index])
  return(models[best_model_index])
  
}


```


Choose best model

```{r}

models = list(
  m1r = m1r,
  m1s = m1s,
  m1t = m1t,
  m1u = m1u
)

select_best_model_of_class(models)

```


## Plot models

Best model: m1r

```{r}

summary(m1r)

data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

m1r_predict <- predictY(m1r, data_predict, var.time = "age_years", draws = TRUE)

m1r_predict <- data.frame(m1r_predict[2], m1r_predict[1])

m1r_predict <- m1r_predict %>%
  rename("mean_m1r" = "pred.Ypred",
         "lower_m1r" = "pred.lower.Ypred",
         "upper_m1r" = "pred.upper.Ypred")

m1r_predict <- m1r_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

knots <- attr(splines::bs(data$age_years, degree = 3, df = 6), "knots")

plot_m1r <- m1r_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)+
  scale_x_continuous(breaks = c(0 , 2, 4, 6, 8, 10, 12, 14, 16, 18))+
  scale_y_continuous(limits = c(12, 25))+
  theme_minimal()

plot_m1r

```

Second best model: m1t

```{r}

summary(m1t)

data_predict <- data.frame(age_years = seq(0,18,length.out=1000))

m1t_predict <- predictY(m1t, data_predict, var.time = "age_years", draws = TRUE)

m1t_predict <- data.frame(m1t_predict[2], m1t_predict[1])

m1t_predict <- m1t_predict %>%
  rename("mean_m1t" = "pred.Ypred",
         "lower_m1t" = "pred.lower.Ypred",
         "upper_m1t" = "pred.upper.Ypred")

m1t_predict <- m1t_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "model"),
               names_sep = "_",
               values_to = "value")

plot_m1t <- m1t_predict %>% 
  ggplot(aes(x = age_years, y = mean, colour = model)) + 
  geom_line()+
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3)+
  geom_vline(xintercept = knots)+
  scale_x_continuous(breaks = c(0 , 2, 4, 6, 8, 10, 12, 14, 16, 18))+
  scale_y_continuous(limits = c(12, 25))+
  theme_minimal()

plot_m1t

```

### Compare best models to expected pattern

```{r}

models_predict <- rbind(m1r_predict, m1t_predict, mloess_predict, mgam_predict, who_predict)


plot_models <- ggplot(aes(x = age_years, y = mean, colour = model), data = models_predict) + 
  geom_line() + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18)) +  
  scale_y_continuous(limits = c(12, 25)) +  
  theme_minimal()

plot_models

```

Models m1r and m1t seem to fit close to the expected patterns from the WHO growth charts. We will continue with the quadratic splines with 2 or 3 manually placed knots. We will estimate the 2-class to 7-class models on ubelix.

