---
title: "Data analysis of association of BMI trajectories and asthma in Leicester Respiratory Cohorts"
author: "Lorenz Leuenberger"
date: "`r date()`"
title-block-banner: "#e4003c"
format: 
  html:
    page-layout: full
    toc: true
    number-sections: false
    code-fold: true
    code-tools: true
    df-print: paged
    embed-resources: true
---

# Working environment and software

## Directory

The folder (Leicester_bmi-trajectories) for the analysis of Leicester Respiratory Cohort data has the following structure:

-   data
    -   raw
    -   processed
    -   interpolated-growth-references
-   R-and-qmd scripts
    -   old
-   output
    -   figures
    -   tables
-   ubelix
    -   batch-files
    -   input
    -   output
    -   r-scripts

I have 6 R-Quarto files for the analysis:

-   00_data-cleaning.qmd: Clean raw data from Leicester Respiratory Cohorts in two parts: i. clean anthropometric data, ii. clean other outcomes and covariates. I already code variables from the raw questionnaire data.
-   01_data-availability.qmd: Check availability of the anthropometric data to choose inclusion and exclusion criteria.
-   02_preparation-for-ubelix.qmd: Use inclusion and exclusion criteria and pseudonymize data.
-   03_best-1-class-model.qmd: Find suitable function (polynomials, splines) to model BMI trajectories.
-   The BMI trajectories were estimated on the high performance computing cluster of the university of Bern (ubelix). The scripts and output is saved in the ubelix folder.
-   04_model-selection_bmi-trajectories.qmd: Selection of optimal BMI trajectory model.
-   05_data-analysis_bmi-trajectories.qmd: Analysis and characterization of BMI trajectories, identify risk groups and risk factors

## Software

Load packages

```{r}
#| output: false
#| warning: false

#Libraries

library(tidyverse)
library(here)
library(gtsummary)
library(cowplot)
library(unibeCols)
library(DescTools)
library(DataExplorer)
library(nnet)
library(Hmisc)
library(lcmm)
library(splines)
library(mice)
library(future)
library(future.apply)
library(ggnewscale)
library(ggtext)
library(survey)
library(srvyr)
library(midoc)
library(dagitty)
library(ggdag)
library(cobalt)

```

Write function to round 0.5 and 1.5 up consistently.

```{r}
#| warning: false

# function to round 0.5 and 1.5 up consistently

round2 = function(x, digits) {
  posneg = sign(x)
  z = abs(x)*10^digits
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^digits
  z*posneg
}



ggtext_size <- function(base_size, ratio = 1) {
  ratio * base_size / ggplot2::.pt
}

```

Function to load BMI trajectory models

```{r}
#| output: false
#| warning: false

load_models_to_list <- function(n_class, group) {
  
  file_path = here("ubelix", "output", "final_splines", paste0(n_class, "-class"), group)
  
  rdata_files <- list.files(path = file_path, full.names = FALSE)
  
  temp_env <- new.env()
  
  for (file in rdata_files) {
    load(file.path(file_path, file), envir = temp_env)
  }
  
  models <- as.list(temp_env)
  rm(temp_env)
  
  models <- models[c(paste0("bmi_m", n_class, "r_", group),
                     paste0("bmi_m", n_class, "s_", group),
                     paste0("bmi_m", n_class, "t_", group),
                     paste0("bmi_m", n_class, "u_", group),
                     paste0("bmi_m", n_class, "v_", group),
                     paste0("bmi_m", n_class, "w_", group))]
  
  return(models)
}

```


# Inclusion and exclusion criteria

Inclusion:

- Include all children with at least 3 BMI measurements, at least 1 during infancy (0-1) and 1 during school age (3-10).

Exclusion:

- Exclude outliers of BMI (recentered z-score \<-5 or \>5)

## Figure 1: Flow diagram of inclusion criteria

Figure S1 was created in Word.

## eTable 1: Data collection in LRC

eTable 1 was created in Word. Adapted from Kuehni et al.

# Data preparation

## Import data

```{r}
#| output: false
#| warning: false

C1_outcomes_covariates <- read.csv(here("data", "processed", "C1_outcomes_covariates.csv"))
C2_outcomes_covariates <- read.csv(here("data", "processed", "C2_outcomes_covariates.csv"))
C3_outcomes_covariates <- read.csv(here("data", "processed", "C3_outcomes_covariates.csv"))

Leicester_anthro_18 <- read_csv(here("data", "processed", "Leicester_anthro_18.csv"))

participant_classmembership <- read_csv(here("data", "processed", "participant_classmembership.csv"))
load(here("data", "processed", "bmi_models_total.Rdata"))
load(here("data", "processed", "bmi_models_female.Rdata"))
load(here("data", "processed", "bmi_models_male.Rdata"))
load(here("data", "processed", "bmi_models_europe.Rdata"))
load(here("data", "processed", "bmi_models_asia.Rdata"))

interpolated_ukwho_bmi <- read.csv(here("data", "interpolated-growth-references", "interpolated_ukwho_bmi.csv"))

```


## Clean growth reference data

Clean growth reference data:

```{r}
#| warning: false

z_score_overweight <- qnorm(0.91)
z_score_obesity <- qnorm(0.98)
z_score_underweight <- qnorm(0.02)

data_bmi_reference <- interpolated_ukwho_bmi %>%
  filter(age_days <= 6574) %>% 
  select(age_days, sex, M, L, S) %>% 
  pivot_wider(id_cols = "age_days", names_from = "sex", values_from = c("M", "L", "S")) %>% 
  mutate(cutoff_overweight_f = M_f * (1 + L_f * S_f * z_score_overweight)^(1 / L_f),
         cutoff_obesity_f = M_f * (1 + L_f * S_f * z_score_obesity)^(1 / L_f),
         cutoff_underweight_f = M_f * (1 + L_f * S_f * z_score_underweight)^(1 / L_f),
         cutoff_overweight_m = M_m * (1 + L_m * S_m * z_score_overweight)^(1 / L_m),
         cutoff_obesity_m = M_m * (1 + L_m * S_m * z_score_obesity)^(1 / L_m),
         cutoff_underweight_m = M_m * (1 + L_m * S_m * z_score_underweight)^(1 / L_m),
         lower_overweight_who = (cutoff_overweight_f+cutoff_overweight_m)/2,
         lower_obesity_who = (cutoff_obesity_f+cutoff_obesity_m)/2,
         lower_underweight_who = -Inf,
         upper_overweight_who = lower_obesity_who,
         upper_obesity_who = Inf,
         upper_underweight_who = (cutoff_underweight_f+cutoff_underweight_m)/2) %>% 
  mutate(age_years = age_days/365.25,
         mean_who = (M_f+M_m)/2) %>% 
  select(age_years, lower_overweight_who, lower_obesity_who, lower_underweight_who, upper_overweight_who, upper_obesity_who, upper_underweight_who, mean_who) %>% 
  pivot_longer(
    cols = c(lower_overweight_who, lower_obesity_who, lower_underweight_who, upper_overweight_who, upper_obesity_who, upper_underweight_who, mean_who),
    names_to = c(".value", "class"),
    names_sep = "_"
  ) %>% 
  mutate(class = case_when(class == "who" ~ "50th percentile",
                           TRUE ~ class)) |> 
  bind_rows(data.frame(age_years = c(0,0,0,0),
                       class = c("underweight", "overweight", "obesity", "50th percentile"),
                       lower = c(-Inf, 15.34684, 16.35499, NA),
                       upper = c(11.01485, 16.35499, Inf, NA),
                       mean = c(NA, NA, NA, 13.54390)))

data_bmi_reference <- data_bmi_reference |> 
  mutate(class = factor(class, levels = c("obesity", "overweight", "50th percentile", "underweight"), labels = c("obesity", "overweight", "50th percentile", "underweight")))

```

## Create variables

### Outcome: BMI trajectories

Create factor with BMI trajectories.

- Trajectory 1: Early overweight resolving
- Trajectory 2: Persistent low BMI
- Trajectory 3: Adolescent onset overweight
- Trajectory 4: Stable normal BMI
- Trajectory 5: Childhood onset Obesity

```{r}
#| warning: false

participant_classmembership <- participant_classmembership %>%
  mutate(class = factor(class, levels = c(1,2,3,4,5), labels = c("Early overweight resolving", "Persistent low BMI", "Adolescent onset overweight", "Stable normal BMI", "Childhood onset obesity")),
         class = relevel(class, ref = 3))

```


### Exposures / Covariates

#### eTable 2: Variable coding

Created in Word.

Clean Leicester covariates data

```{r}
#| warning: false

C1_covariates <- C1_outcomes_covariates %>% 
  select(serialno, cohort_number,
         birth_year, sex, gestational_age, birth_weight, older_siblings, age_mother_delivery, smoking_mother_pregnancy, breast_feeding, breast_feeding_duration, day_care,
         ethnicity_cohort, ethnicity, ethnicity_mother, ethnicity_father,
         age_mother_education, age_father_education,
         townsend_2001, townsend_2001_quintile, urban_rural,
         age_questionnaire_90, age_questionnaire_92, age_questionnaire_98, age_questionnaire_03, age_questionnaire_10, age_questionnaire_12,
         smoking_mother_90, smoking_mother_92, smoking_mother_98, smoking_mother_03, smoking_mother_10, smoking_mother_12, smoking_others_90, smoking_others_92, smoking_others_98, smoking_others_03, smoking_others_10, smoking_others_12,
         smoking_personal_92, smoking_personal_98, smoking_personal_03, smoking_personal_10, smoking_personal_12,
         physical_activity_hrs_per_day_90, physical_activity_hrs_per_day_92, physical_activity_hrs_per_day_98, physical_activity_vigorous_hrs_per_week_03, physical_activity_vigorous_hrs_per_week_10, physical_activity_vigorous_hrs_per_week_12,
         fresh_fruit_per_week_03, fresh_fruit_per_week_10, vegetables_per_week_10, crisps_per_week_10, chocolate_per_week_10)

C2_covariates <- C2_outcomes_covariates %>%
  mutate(age_questionnaire_05 = case_when(is.na(age_questionnaire_children_05) ~ age_questionnaire_parents_05,
                                          TRUE ~ age_questionnaire_children_05),
         age_questionnaire_ML = case_when(is.na(age_questionnaire_children_ML) ~ age_questionnaire_parents_ML,
                                          TRUE ~ age_questionnaire_children_ML)) |> 
  select(serialno, cohort_number,
         birth_year, sex, gestational_age, birth_weight, older_siblings, age_mother_delivery, smoking_mother_pregnancy, breast_feeding, breast_feeding_duration, day_care,
         ethnicity_cohort, ethnicity, ethnicity_mother, ethnicity_father,
         age_mother_education, age_father_education,
         townsend_2001, townsend_2001_quintile, urban_rural,
         age_questionnaire_98, age_questionnaire_01, age_questionnaire_03, age_questionnaire_05, age_questionnaire_06, age_questionnaire_ML, age_questionnaire_He1, age_questionnaire_He2, age_questionnaire_10, age_questionnaire_12,
         smoking_mother_98, smoking_mother_01, smoking_mother_03, smoking_mother_05, smoking_mother_06, smoking_mother_ML, smoking_mother_He1, smoking_mother_He2, smoking_mother_10, smoking_mother_12, smoking_others_98, smoking_others_01, smoking_others_03, smoking_others_05, smoking_others_06, smoking_others_ML, smoking_others_He1, smoking_others_He2, smoking_others_10, smoking_others_12,
         smoking_personal_98, smoking_personal_01, smoking_personal_05, smoking_personal_06, smoking_personal_children_ML, smoking_personal_ever_children_ML, smoking_personal_ever_3_months_children_ML, smoking_personal_ML, smoking_personal_He1, smoking_personal_ever_He1, smoking_personal_ever_3_months_He1, smoking_personal_He2, smoking_personal_ever_He2,
         physical_activity_hrs_per_day_98, physical_activity_hrs_per_day_01, physical_activity_hrs_per_day_03, physical_activity_hrs_per_day_05, physical_activity_hrs_per_day_06, physical_activity_hrs_per_day_ML, physical_activity_vigorous_hrs_per_week_03, physical_activity_vigorous_hrs_per_week_05, physical_activity_vigorous_hrs_per_week_06, physical_activity_vigorous_hrs_per_week_ML, physical_activity_vigorous_hrs_per_week_He1, physical_activity_vigorous_hrs_per_week_He2, physical_activity_vigorous_hrs_per_week_10, physical_activity_vigorous_hrs_per_week_12,
         fresh_fruit_per_week_01, fresh_fruit_per_week_03, fresh_fruit_per_week_05, fresh_fruit_per_week_06, fresh_fruit_per_week_He1, fresh_fruit_per_week_He2, fresh_fruit_per_week_ML, fresh_fruit_per_week_10, vegetables_per_week_10, crisps_per_week_05, crisps_per_week_06, crisps_per_week_ML, crisps_per_week_10, chocolate_per_week_05, chocolate_per_week_06, chocolate_per_week_ML, chocolate_per_week_10, diet_type_01, diet_type_03, diet_type_05, diet_type_06, diet_type_ML)

C3_covariates <- C3_outcomes_covariates %>%
  mutate(age_questionnaire_05 = case_when(is.na(age_questionnaire_children_05) ~ age_questionnaire_parents_05,
                                          TRUE ~ age_questionnaire_children_05),
         age_questionnaire_ML = case_when(is.na(age_questionnaire_children_ML) ~ age_questionnaire_parents_ML,
                                          TRUE ~ age_questionnaire_children_ML)) |> 
  select(serialno, cohort_number,
         birth_year, sex, gestational_age, birth_weight, older_siblings, age_mother_delivery, smoking_mother_pregnancy, breast_feeding, breast_feeding_duration, day_care,
         ethnicity_cohort, ethnicity, ethnicity_mother, ethnicity_father,
         age_mother_education, age_father_education,
         townsend_2001, townsend_2001_quintile, urban_rural,
         age_questionnaire_98, age_questionnaire_99, age_questionnaire_01, age_questionnaire_03, age_questionnaire_05, age_questionnaire_06, age_questionnaire_ML, age_questionnaire_He1, age_questionnaire_He2, age_questionnaire_10, age_questionnaire_12,
         smoking_mother_98, smoking_mother_99, smoking_mother_01, smoking_mother_03, smoking_mother_05, smoking_mother_06, smoking_mother_ML, smoking_mother_He1, smoking_mother_He2, smoking_mother_10, smoking_mother_12, smoking_others_98, smoking_others_99, smoking_others_01, smoking_others_03, smoking_others_05, smoking_others_06, smoking_others_ML, smoking_others_He1, smoking_others_He2, smoking_others_10, smoking_others_12,
         smoking_personal_98, smoking_personal_01, smoking_personal_05, smoking_personal_06, smoking_personal_children_ML, smoking_personal_ever_children_ML, smoking_personal_ever_3_months_children_ML, smoking_personal_ML, smoking_personal_He1, smoking_personal_ever_He1, smoking_personal_ever_3_months_He1, smoking_personal_He2, smoking_personal_ever_He2, smoking_personal_ever_3_months_He2, smoking_personal_10, smoking_personal_ever_1_year_10, smoking_personal_12, smoking_personal_ever_1_year_12,
         physical_activity_hrs_per_day_98, physical_activity_hrs_per_day_99, physical_activity_hrs_per_day_01, physical_activity_hrs_per_day_03, physical_activity_hrs_per_day_05, physical_activity_hrs_per_day_06, physical_activity_hrs_per_day_ML, physical_activity_vigorous_hrs_per_week_03, physical_activity_vigorous_hrs_per_week_05, physical_activity_vigorous_hrs_per_week_06, physical_activity_vigorous_hrs_per_week_ML, physical_activity_vigorous_hrs_per_week_He1, physical_activity_vigorous_hrs_per_week_He2, physical_activity_vigorous_hrs_per_week_10, physical_activity_vigorous_hrs_per_week_12,
         fresh_fruit_per_week_01, fresh_fruit_per_week_03, fresh_fruit_per_week_05, fresh_fruit_per_week_06, fresh_fruit_per_week_He1, fresh_fruit_per_week_He2, fresh_fruit_per_week_ML, fresh_fruit_per_week_10, vegetables_per_week_10, crisps_per_week_05, crisps_per_week_06, crisps_per_week_ML, crisps_per_week_10, chocolate_per_week_05, chocolate_per_week_06, chocolate_per_week_ML, chocolate_per_week_10, diet_type_01, diet_type_03, diet_type_05, diet_type_06, diet_type_ML)

```

#### Combine cohorts

Combine cohorts

```{r}
#| warning: false

Leicester_covariates <- bind_rows(C1_covariates, C2_covariates, C3_covariates)

Leicester_data <- Leicester_covariates %>% 
  left_join(participant_classmembership, by = "serialno")

Leicester_data <- Leicester_data |> 
  mutate(class = factor(class,
                        levels = c("Stable normal BMI", "Persistent low BMI", "Early overweight resolving", "Childhood onset obesity", "Adolescent onset overweight")))

```

#### Code variables

```{r}
#| warning: false

Leicester_data <- Leicester_data |> 
  mutate(cohort_number = factor(cohort_number),
         cohort_number = fct_relevel(cohort_number, c("Cohort 1 (1990)", "Cohort 2 (1998a)", "Cohort 3 (1998b)")),
         birth_year_cat = factor(case_when(birth_year < 1990 ~ "1985-1989",
                                             TRUE ~ "1993-1997"),
                                   levels = c("1985-1989", "1993-1997")), 
         ethnicity = factor(ethnicity),
         ethnicity_cohort = factor(ethnicity_cohort, levels = c("whites", "asians"), labels = c("White", "South Asian")),
         smoking_mother_pregnancy = factor(smoking_mother_pregnancy),
         smoking_mother_pregnancy = fct_relevel(smoking_mother_pregnancy, c("no", "yes")),
         breast_feeding = factor(breast_feeding),
         breast_feeding = fct_relevel(breast_feeding, c("no", "yes")),
         breast_feeding_duration = factor(breast_feeding_duration),
         breast_feeding_duration = fct_relevel(breast_feeding_duration, c("0 months", "<=1 month", "1-3 months", "4-6 months", ">=6 months")))

Leicester_data <- Leicester_data %>% 
  mutate(ethnic = factor(case_when(ethnicity == "European" ~ "European",
                                   ethnicity == "Asian" ~ "Asian",
                                   TRUE ~ "Mixed/Other")),
         ethnic = fct_relevel(ethnic, c("European", "Asian", "Mixed/Other")))

# urbanization

Leicester_data <- Leicester_data |> 
  mutate(urban_rural = factor(urban_rural, levels = c("urban", "rural")))

# townsend

Leicester_data <- Leicester_data |> 
  mutate(townsend_2001_quintile = factor(townsend_2001_quintile, levels = c(1,2,3,4,5)))

# education

Leicester_data <- Leicester_data |> 
  rowwise() |> 
  mutate(age_parental_education = case_when(is.na(age_mother_education) & is.na(age_father_education) ~ NA,
                                            TRUE ~ max(c(age_mother_education, age_father_education), na.rm = TRUE)),
         parental_education_cat = factor(case_when(age_father_education >= 20 | age_mother_education >= 20 ~ "higher education",
                                              age_father_education >= 17 | age_mother_education >= 17 ~ "further education",
                                              age_father_education < 17 | age_mother_education < 17 ~ "compulsory education",
                                              TRUE ~ NA),
                                         levels = c("compulsory education", "further education", "higher education"))) |> 
  ungroup()


# sex

Leicester_data <- Leicester_data |> 
  mutate(sex = factor(sex, levels = c("male", "female")))

# Birth weight

Leicester_data <- Leicester_data |> 
  mutate(low_birth_weight = case_when(is.na(birth_weight) ~ NA,
                                      birth_weight < 2500 ~ "Low",
                                      TRUE ~ "Normal"))|> 
  mutate(low_birth_weight = factor(low_birth_weight, levels = c("Normal", "Low"), labels = c("Normal", "Low")))

# gestational age

Leicester_data <- Leicester_data |> 
  mutate(gestational_age = case_when(gestational_age == 0 ~ NA,
                                     TRUE ~ gestational_age)) |> 
  mutate(term_birth = case_when(gestational_age < 37 ~ "preterm",
                                gestational_age >= 37 ~ "term",
                                TRUE ~ NA))

Leicester_data |> 
  select(gestational_age, term_birth) |> 
  tbl_summary()

# older siblings

Leicester_data <- Leicester_data |> 
  mutate(number_older_siblings = older_siblings,
         older_siblings = case_when(number_older_siblings == 0 ~ "no",
                                    number_older_siblings > 0 ~ "yes",
                                    TRUE ~ NA),
         birth_order = number_older_siblings + 1,
         birth_order_cat = factor(case_when(birth_order == 1 ~ "first child",
                                            birth_order == 2 ~ "second child",
                                            birth_order >= 3 ~ "third or later child",
                                            TRUE ~ NA)))

#smoking mother pregnancy

Leicester_data |> select(smoking_mother_pregnancy, cohort_number) |> tbl_summary(by = cohort_number)
Leicester_data |> select(smoking_mother_pregnancy, ethnicity_cohort) |> tbl_summary(by = ethnicity_cohort)

# smoking others


Leicester_data |> select(smoking_mother_90, smoking_mother_92, smoking_mother_98, smoking_mother_99, smoking_mother_01, smoking_mother_03, smoking_mother_05, smoking_mother_06, smoking_mother_ML, smoking_mother_He1, smoking_mother_He2, smoking_mother_10, smoking_mother_12, ethnicity_cohort) |> 
  tbl_summary(by = ethnicity_cohort,
              type = all_dichotomous() ~ "categorical")

Leicester_data |> select(smoking_others_90, smoking_others_92, smoking_others_98, smoking_others_99, smoking_others_01, smoking_others_03, smoking_others_05, smoking_others_06, smoking_others_ML, smoking_others_He1, smoking_others_He2, smoking_others_10, smoking_others_12, ethnicity_cohort) |> 
  tbl_summary(by = ethnicity_cohort,
              type = all_dichotomous() ~ "categorical")


Leicester_data <- Leicester_data |> 
  mutate(smoking_others_ever = factor(case_when(smoking_mother_90 == "yes" ~ "yes",
                                                smoking_mother_92 == "yes" ~ "yes",
                                                smoking_mother_98 == "yes" ~ "yes",
                                                smoking_mother_99 == "yes" ~ "yes",
                                                smoking_mother_01 == "yes" ~ "yes",
                                                smoking_mother_03 == "yes" ~ "yes",
                                                smoking_mother_05 == "yes" ~ "yes",
                                                smoking_mother_06 == "yes" ~ "yes",
                                                smoking_mother_ML == "yes" ~ "yes",
                                                smoking_mother_He1 == "yes" ~ "yes",
                                                smoking_mother_He2 == "yes" ~ "yes",
                                                smoking_mother_10 == "yes" ~ "yes",
                                                smoking_mother_12 == "yes" ~ "yes",
                                                smoking_others_90 == "yes" ~ "yes",
                                                smoking_others_92 == "yes" ~ "yes",
                                                smoking_others_98 == "yes" ~ "yes",
                                                smoking_others_99 == "yes" ~ "yes",
                                                smoking_others_01 == "yes" ~ "yes",
                                                smoking_others_03 == "yes" ~ "yes",
                                                smoking_others_05 == "yes" ~ "yes",
                                                smoking_others_06 == "yes" ~ "yes",
                                                smoking_others_ML == "yes" ~ "yes",
                                                smoking_others_He1 == "yes" ~ "yes",
                                                smoking_others_He2 == "yes" ~ "yes",
                                                smoking_others_10 == "yes" ~ "yes",
                                                smoking_others_12 == "yes" ~ "yes",
                                                smoking_mother_90 == "no" ~ "no",
                                                smoking_mother_92 == "no" ~ "no",
                                                smoking_mother_98 == "no" ~ "no",
                                                smoking_mother_99 == "no" ~ "no",
                                                smoking_mother_01 == "no" ~ "no",
                                                smoking_mother_03 == "no" ~ "no",
                                                smoking_mother_05 == "no" ~ "no",
                                                smoking_mother_06 == "no" ~ "no",
                                                smoking_mother_ML == "no" ~ "no",
                                                smoking_mother_He1 == "no" ~ "no",
                                                smoking_mother_He2 == "no" ~ "no",
                                                smoking_mother_10 == "no" ~ "no",
                                                smoking_mother_12 == "no" ~ "no",
                                                smoking_others_90 == "no" ~ "no",
                                                smoking_others_92 == "no" ~ "no",
                                                smoking_others_98 == "no" ~ "no",
                                                smoking_others_99 == "no" ~ "no",
                                                smoking_others_01 == "no" ~ "no",
                                                smoking_others_03 == "no" ~ "no",
                                                smoking_others_05 == "no" ~ "no",
                                                smoking_others_06 == "no" ~ "no",
                                                smoking_others_ML == "no" ~ "no",
                                                smoking_others_He1 == "no" ~ "no",
                                                smoking_others_He2 == "no" ~ "no",
                                                smoking_others_10 == "no" ~ "no",
                                                smoking_others_12 == "no" ~ "no",
                                                
                                                TRUE ~ NA)),
         smoking_others_ever = fct_relevel(smoking_others_ever, c("no", "yes")))
  
Leicester_data |> select(smoking_others_ever, ethnicity_cohort) |> 
  tbl_summary(by = "ethnicity_cohort")


#smoking personal

Leicester_data <- Leicester_data |> 
  mutate(smoking_personal_ever = case_when(smoking_personal_92 == "yes" ~ "yes",
                                           smoking_personal_98 == "yes" ~ "yes",
                                           smoking_personal_01 == "yes" ~ "yes",
                                           smoking_personal_03 == "yes" ~ "yes",
                                           smoking_personal_05 == "yes" ~ "yes",
                                           smoking_personal_06 == "yes" ~ "yes",
                                           smoking_personal_ML == "yes" ~ "yes",
                                           smoking_personal_children_ML == "yes" ~ "yes",
                                           smoking_personal_ever_children_ML == "yes" ~ "yes",
                                           smoking_personal_ever_3_months_children_ML == "yes" ~ "yes",
                                           smoking_personal_He1 == "yes" ~ "yes",
                                           smoking_personal_ever_He1 == "yes" ~ "yes",
                                           smoking_personal_ever_3_months_He1 == "yes" ~ "yes",
                                           smoking_personal_He2 == "yes" ~ "yes",
                                           smoking_personal_ever_He2 == "yes" ~ "yes",
                                           smoking_personal_ever_3_months_He2 == "yes" ~ "yes",
                                           smoking_personal_10 == "yes" ~ "yes",
                                           smoking_personal_ever_1_year_10 == "yes" ~ "yes",
                                           smoking_personal_12 == "yes" ~ "yes",
                                           smoking_personal_ever_1_year_12 == "yes" ~ "yes",
                                           smoking_personal_92 == "no" ~ "no",
                                           smoking_personal_98 == "no" ~ "no",
                                           smoking_personal_01 == "no" ~ "no",
                                           smoking_personal_03 == "no" ~ "no",
                                           smoking_personal_05 == "no" ~ "no",
                                           smoking_personal_06 == "no" ~ "no",
                                           smoking_personal_ML == "no" ~ "no",
                                           smoking_personal_children_ML == "no" ~ "no",
                                           smoking_personal_ever_children_ML == "no" ~ "no",
                                           smoking_personal_ever_3_months_children_ML == "no" ~ "no",
                                           smoking_personal_He1 == "no" ~ "no",
                                           smoking_personal_ever_He1 == "no" ~ "no",
                                           smoking_personal_ever_3_months_He1 == "no" ~ "no",
                                           smoking_personal_He2 == "no" ~ "no",
                                           smoking_personal_ever_He2 == "no" ~ "no",
                                           smoking_personal_ever_3_months_He2 == "no" ~ "no",
                                           smoking_personal_10 == "no" ~ "no",
                                           smoking_personal_ever_1_year_10 == "no" ~ "no",
                                           smoking_personal_12 == "no" ~ "no",
                                           smoking_personal_ever_1_year_12 == "no" ~ "no",
                                           TRUE ~ NA))

Leicester_data |> select(smoking_personal_ever) |> 
  tbl_summary(type = all_dichotomous() ~ "categorical",
              missing_stat = "{N_miss} ({p_miss}%)")

# physical activity

Leicester_data |> 
  select(serialno,
         physical_activity_hrs_per_day_90, physical_activity_hrs_per_day_92, physical_activity_hrs_per_day_98, physical_activity_hrs_per_day_99, physical_activity_hrs_per_day_01, physical_activity_hrs_per_day_03, physical_activity_hrs_per_day_05, physical_activity_hrs_per_day_06, physical_activity_hrs_per_day_ML,
         age_questionnaire_90, age_questionnaire_92, age_questionnaire_98, age_questionnaire_99, age_questionnaire_01, age_questionnaire_03, age_questionnaire_05, age_questionnaire_06, age_questionnaire_ML) |> 
    pivot_longer(
    cols = -serialno,
    names_to = c("measure", "year"),
    names_pattern = "(.*)_(\\d{2}|ML)",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = measure,
    values_from = value
  ) |> 
  ggplot(aes(x = age_questionnaire, y = physical_activity_hrs_per_day))+
  geom_point()

Leicester_data |> 
  select(serialno,
         physical_activity_vigorous_hrs_per_week_03, physical_activity_vigorous_hrs_per_week_05, physical_activity_vigorous_hrs_per_week_06, physical_activity_vigorous_hrs_per_week_ML, physical_activity_vigorous_hrs_per_week_He1, physical_activity_vigorous_hrs_per_week_He2, physical_activity_vigorous_hrs_per_week_10, physical_activity_vigorous_hrs_per_week_12,
         age_questionnaire_03, age_questionnaire_05, age_questionnaire_06, age_questionnaire_ML, age_questionnaire_He1, age_questionnaire_He2, age_questionnaire_10, age_questionnaire_12) |>
  pivot_longer(
    cols = -serialno,
    names_to = c("measure", "year"),
    names_pattern = "(.*)_(\\d{2}|ML|He1|He2)",
    values_to = "value"
  ) |> 
    pivot_wider(
    names_from = measure,
    values_from = value
  ) |> 
  ggplot(aes(x = age_questionnaire, y = physical_activity_vigorous_hrs_per_week))+
  geom_point()


Leicester_data <- Leicester_data |>
  rowwise() |> 
  mutate(mean_physical_activity_hrs_per_day_under_5 = mean(c(
    ifelse(age_questionnaire_90 < 5, physical_activity_hrs_per_day_90, NA),
    ifelse(age_questionnaire_92 < 5, physical_activity_hrs_per_day_92, NA),
    ifelse(age_questionnaire_98 < 5, physical_activity_hrs_per_day_98, NA),
    ifelse(age_questionnaire_99 < 5, physical_activity_hrs_per_day_99, NA),
    ifelse(age_questionnaire_01 < 5, physical_activity_hrs_per_day_01, NA),
    ifelse(age_questionnaire_03 < 5, physical_activity_hrs_per_day_03, NA),
    ifelse(age_questionnaire_05 < 5, physical_activity_hrs_per_day_05, NA),
    ifelse(age_questionnaire_06 < 5, physical_activity_hrs_per_day_06, NA),
    ifelse(age_questionnaire_ML < 5, physical_activity_hrs_per_day_ML, NA)), na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(mean_physical_activity_hrs_per_day_under_5 = case_when(mean_physical_activity_hrs_per_day_under_5 == "NaN" ~ NA,
                                                                TRUE ~ as.numeric(mean_physical_activity_hrs_per_day_under_5))) |> 
  mutate(physical_activity_recommended_under_5 = case_when(mean_physical_activity_hrs_per_day_under_5 < 3 ~ "not reached",
                                                   mean_physical_activity_hrs_per_day_under_5 >= 3 ~ "reached"))


Leicester_data |> 
  select(mean_physical_activity_hrs_per_day_under_5, physical_activity_recommended_under_5) |> 
  tbl_summary()

Leicester_data <- Leicester_data |>
  rowwise() |> 
  mutate(mean_physical_activity_hrs_per_week_5_to_18 = mean(c(
    ifelse(age_questionnaire_03 > 5 & age_questionnaire_03 <= 18, physical_activity_vigorous_hrs_per_week_03, NA),
    ifelse(age_questionnaire_05 > 5 & age_questionnaire_05 <= 18, physical_activity_vigorous_hrs_per_week_05, NA),
    ifelse(age_questionnaire_06 > 5 & age_questionnaire_06 <= 18, physical_activity_vigorous_hrs_per_week_06, NA),
    ifelse(age_questionnaire_ML > 5 & age_questionnaire_ML <= 18, physical_activity_vigorous_hrs_per_week_ML, NA),
    ifelse(age_questionnaire_He1 > 5 & age_questionnaire_He1 <= 18, physical_activity_vigorous_hrs_per_week_He1, NA),
    ifelse(age_questionnaire_He2 > 5 & age_questionnaire_He2 <= 18, physical_activity_vigorous_hrs_per_week_He2, NA),
    ifelse(age_questionnaire_10 > 5 & age_questionnaire_10 <= 18, physical_activity_vigorous_hrs_per_week_10, NA),
    ifelse(age_questionnaire_12 > 5 & age_questionnaire_12 <= 18, physical_activity_vigorous_hrs_per_week_12, NA)), na.rm = TRUE)) |>
  ungroup() |> 
  mutate(mean_physical_activity_hrs_per_week_5_to_18 = case_when(mean_physical_activity_hrs_per_week_5_to_18 == "NaN" ~ NA,
                                                                TRUE ~ as.numeric(mean_physical_activity_hrs_per_week_5_to_18))) |> 
  mutate(physical_activity_recommended_5_to_18 = case_when(mean_physical_activity_hrs_per_week_5_to_18 < 7 ~ "not reached",
                                                   mean_physical_activity_hrs_per_week_5_to_18 >= 7 ~ "reached",
                                                   TRUE ~ NA))


Leicester_data <- Leicester_data |> 
  mutate(physical_activity_cat_mean = factor(case_when(mean_physical_activity_hrs_per_week_5_to_18 >= 6 ~ ">=6 hrs per week",
                                             mean_physical_activity_hrs_per_week_5_to_18 >= 4 ~ "4-5 hrs per week",
                                             mean_physical_activity_hrs_per_week_5_to_18 >= 2 ~ "2-3 hrs per week",
                                             mean_physical_activity_hrs_per_week_5_to_18 < 2 ~ "0-1 hr per week"),
                                        levels = c("0-1 hr per week", "2-3 hrs per week", "4-5 hrs per week", ">=6 hrs per week")))


Leicester_data <- Leicester_data |> 
  mutate(physical_activity_cat_minimal = factor(case_when(mean_physical_activity_hrs_per_week_5_to_18 >= 4 ~ ">= 4 hrs per week",
                                                          mean_physical_activity_hrs_per_week_5_to_18 < 4 ~ "0-3 hrs per week"),
                                                levels = c("0-3 hrs per week", ">= 4 hrs per week")))


Leicester_data <- Leicester_data |> 
  mutate(physical_activity_cat13_18 = factor(case_when(birth_year_cat == "1985-1989" & physical_activity_vigorous_hrs_per_week_03 >=6 ~ ">=6 hrs per week",
                                                 birth_year_cat == "1985-1989" & physical_activity_vigorous_hrs_per_week_03 >= 4 ~ "4-5 hrs per week",
                                                 birth_year_cat == "1985-1989" & physical_activity_vigorous_hrs_per_week_03 >= 2 ~ "2-3 hrs per week",
                                                 birth_year_cat == "1985-1989" & physical_activity_vigorous_hrs_per_week_03 <2 ~ "0-1 hr per week",
                                                 
                                                 birth_year_cat == "1993-1997" & physical_activity_vigorous_hrs_per_week_10 >=6 ~ ">=6 hrs per week",
                                                 birth_year_cat == "1993-1997" & physical_activity_vigorous_hrs_per_week_10 >= 4 ~ "4-5 hrs per week",
                                                 birth_year_cat == "1993-1997" & physical_activity_vigorous_hrs_per_week_10 >= 2 ~ "2-3 hrs per week",
                                                 birth_year_cat == "1993-1997" & physical_activity_vigorous_hrs_per_week_10 <2 ~ "0-1 hr per week"),
                                       levels = c("0-1 hr per week", "2-3 hrs per week", "4-5 hrs per week", ">=6 hrs per week")))

Leicester_data |> 
  select(mean_physical_activity_hrs_per_week_5_to_18, physical_activity_recommended_5_to_18, physical_activity_cat_mean, physical_activity_cat13_18) |> 
  tbl_summary()




# diet

Leicester_data <- Leicester_data |> 
  mutate(diet = case_when(diet_type_01 == "Mostly English with Asian" | diet_type_03 == "Mostly English with Asian" | diet_type_05 == "Mostly English with Asian" | diet_type_06 == "Mostly English with Asian" | diet_type_ML == "Mostly English with Asian" ~ "English and Asian",
                          (diet_type_01 == "Only English" | diet_type_03 == "Only English" | diet_type_05 == "Only English" | diet_type_06 == "Only English" | diet_type_ML == "Only English") & (diet_type_01 == "Only Asian" | diet_type_03 == "Only Asian" | diet_type_05 == "Only Asian" | diet_type_06 == "Only Asian" | diet_type_ML == "Only Asian" | diet_type_01 == "Mostly Asian" | diet_type_03 == "Mostly Asian" | diet_type_05 == "Mostly Asian" | diet_type_06 == "Mostly Asian" | diet_type_ML == "Mostly Asian") ~ "English and Asian",
                          diet_type_01 == "Only English" | diet_type_03 == "Only English" | diet_type_05 == "Only English" | diet_type_06 == "Only English" | diet_type_ML == "Only English" ~ "English",
                          diet_type_01 == "Only Asian" | diet_type_03 == "Only Asian" | diet_type_05 == "Only Asian" | diet_type_06 == "Only Asian" | diet_type_ML == "Only Asian" | diet_type_01 == "Mostly Asian" | diet_type_03 == "Mostly Asian" | diet_type_05 == "Mostly Asian" | diet_type_06 == "Mostly Asian" | diet_type_ML == "Mostly Asian" ~ "Asian",
                          TRUE ~ NA),
         diet = factor(diet, levels = c("English", "English and Asian", "Asian")))

Leicester_data <- Leicester_data |> 
  rowwise() |> 
  mutate(mean_fresh_fruit_per_week = case_when(birth_year_cat == "1985-1989" ~ fresh_fruit_per_week_03,
                                               birth_year_cat == "1993-1997" ~ mean(c(fresh_fruit_per_week_01, fresh_fruit_per_week_03, fresh_fruit_per_week_05, fresh_fruit_per_week_06, fresh_fruit_per_week_ML, fresh_fruit_per_week_He1, fresh_fruit_per_week_He2, fresh_fruit_per_week_10), na.rm = TRUE))) |> 
  ungroup()

Leicester_data <- Leicester_data |> 
  mutate(fresh_fruit_cat = factor(case_when(mean_fresh_fruit_per_week >=10.5 ~ "twice per day",
                                       mean_fresh_fruit_per_week >= 5 ~ "once daily",
                                       mean_fresh_fruit_per_week > 1 ~ "2-4 per week",
                                       mean_fresh_fruit_per_week <= 1 ~ "0-1 per week"),
                                  levels = c("0-1 per week", "2-4 per week", "once daily", "twice per day")),
         fresh_fruit_cat_mean = factor(case_when(mean_fresh_fruit_per_week >= 5 ~ "Once daily or more",
                                                 mean_fresh_fruit_per_week < 5 ~ "less than once daily"),
                                       levels = c("Once daily or more", "less than once daily")))


Leicester_data |> 
  select(mean_fresh_fruit_per_week, fresh_fruit_cat, fresh_fruit_cat_mean) |> 
  tbl_summary()

Leicester_data <- Leicester_data |> 
  rowwise() |> 
  mutate(mean_vegetables_per_week = mean(c(vegetables_per_week_10), na.rm = TRUE)) |> 
  ungroup()

Leicester_data <- Leicester_data |> 
  mutate(vegetables_cat = factor(case_when(mean_vegetables_per_week >=10.5 ~ "twice per week",
                                       mean_vegetables_per_week >= 5 ~ "once daily",
                                       mean_vegetables_per_week > 1 ~ "2-4 per week",
                                       mean_vegetables_per_week <= 1 ~ "0-1 per week")),
         vegetables_cat_mean = factor(case_when(mean_vegetables_per_week >= 5 ~ "Once daily or more",
                                                 mean_vegetables_per_week < 5 ~ "less than once daily"),
                                       levels = c("Once daily or more", "less than once daily")))


Leicester_data |> 
  select(mean_vegetables_per_week, vegetables_cat, vegetables_cat_mean) |> 
  tbl_summary()

Leicester_data <- Leicester_data |> 
  rowwise() |> 
  mutate(mean_crisps_per_week = mean(c(crisps_per_week_05, crisps_per_week_06, crisps_per_week_ML, crisps_per_week_10), na.rm = TRUE)) |> 
  ungroup()

Leicester_data <- Leicester_data |> 
  mutate(crisps_cat = factor(case_when(mean_crisps_per_week >=10.5 ~ "twice per week",
                                       mean_crisps_per_week >= 5 ~ "once daily",
                                       mean_crisps_per_week > 1 ~ "2-4 per week",
                                       mean_crisps_per_week <= 1 ~ "0-1 per week")),
         crisps_cat_mean = factor(case_when(mean_crisps_per_week >= 5 ~ "Once daily or more",
                                                 mean_crisps_per_week < 5 ~ "less than once daily"),
                                       levels = c("Once daily or more", "less than once daily")))


Leicester_data |> 
  select(mean_crisps_per_week, crisps_cat, crisps_cat_mean) |> 
  tbl_summary()

Leicester_data <- Leicester_data |> 
  rowwise() |> 
  mutate(mean_chocolate_per_week = mean(c(chocolate_per_week_05, chocolate_per_week_06, chocolate_per_week_ML, chocolate_per_week_10), na.rm = TRUE)) |> 
  ungroup()

Leicester_data <- Leicester_data |> 
  mutate(chocolate_cat = factor(case_when(mean_chocolate_per_week >=10.5 ~ "twice per week",
                                       mean_chocolate_per_week >= 5 ~ "once daily",
                                       mean_chocolate_per_week > 1 ~ "2-4 per week",
                                       mean_chocolate_per_week <= 1 ~ "0-1 per week")),
         chocolate_cat_mean = factor(case_when(mean_chocolate_per_week >= 5 ~ "Once daily or more",
                                                 mean_chocolate_per_week < 5 ~ "less than once daily"),
                                       levels = c("Once daily or more", "less than once daily")))


Leicester_data |> 
  select(mean_chocolate_per_week, chocolate_cat, chocolate_cat_mean) |> 
  tbl_summary()

```

## Inclusion/Exclusion

```{r}

Leicester_data <- Leicester_data |> 
  mutate(inclusion = case_when(is.na(class) ~ "excluded",
                               TRUE ~ "included"),
         inclusion = factor(inclusion, levels = c( "excluded", "included")))

```


# Descriptive information

## Outcome

### Number of measurements per participant and by age group

### eTable 5: Number of measurements per participant

In hlme models of the lcmm package, participants with more BMI measurements contribute more information towards the BMI trajectories.
Hence, I want to check the distribution of measurements per participant

```{r}

Leicester_anthro_18 |> 
  count(serialno, name = "number_bmi_measurements") |> 
  ggplot(aes(x = number_bmi_measurements)) +
  geom_histogram()+
  scale_x_continuous(breaks = c(3,6,9,12))+
  theme_minimal()


Leicester_anthro_18 |> 
  count(serialno, name = "number_bmi_measurements") |> 
  select(number_bmi_measurements) |> 
  tbl_summary()

Leicester_anthro_18 |> 
  count(serialno, name = "number_bmi_measurements") |> 
  select(number_bmi_measurements) |> 
  tbl_summary(type = everything() ~ "categorical")

```

Most of the participants have 3-5 measurements. There are very few with 10 or more measurements.

### eTable 6: Number of measurements by age group

```{r}

Leicester_anthro_18 |> 
  mutate(age = trunc(age_years)) |> 
  count(age, name = "number_bmi_measurements") |> 
  mutate(sum = sum(number_bmi_measurements),
         perc = number_bmi_measurements/sum)



```

Number of participants with a measurements in first months of life.

```{r}

Leicester_data |> 
  filter(inclusion == "included") |> 
  left_join(Leicester_anthro_18 |> mutate(BMI_first_months = age_days/365.25*12 < 8) |> group_by(serialno) |> arrange(desc(BMI_first_months)) |> slice(1) |> ungroup() |> select(serialno, BMI_first_months), by = "serialno") |> 
  select(BMI_first_months) |> 
  tbl_summary()

```


### BMI trajectories

BMI Trajectories by cohort.

```{r}
#| warning: false

Leicester_data %>% 
  select(cohort_number, class) %>% 
  tbl_summary(by = "cohort_number",
              type = all_dichotomous() ~ "categorical") %>% 
  add_overall()


Leicester_data %>% 
  filter(inclusion == "included") |> 
  select(class) |> 
  tbl_summary(type = all_dichotomous() ~ "categorical",
              missing = "no")

```


## Exposures / Covariates

```{r}
#| warning: false


plot_histogram(Leicester_data |> select(birth_year, gestational_age, birth_weight, age_mother_delivery, birth_order, age_mother_education, age_father_education, age_parental_education, townsend_2001, mean_physical_activity_hrs_per_week_5_to_18, mean_fresh_fruit_per_week))

plot_bar(Leicester_data |> select(birth_year_cat, sex, ethnicity, ethnic, townsend_2001_quintile, urban_rural, parental_education_cat, birth_order_cat, breast_feeding, breast_feeding_duration, smoking_mother_pregnancy, smoking_others_ever, smoking_personal_ever, urban_rural, physical_activity_recommended_under_5, physical_activity_recommended_5_to_18, physical_activity_cat_mean, fresh_fruit_cat_mean))

```

Check townsend quintiles

```{r}

Leicester_data |> 
  select(townsend_2001, townsend_2001_quintile) |> 
  tbl_summary(by = townsend_2001_quintile,
              statistic = all_continuous() ~ "{min}, {max}")

```



Characteristics of the study participants.

```{r}
#| warning: false

Leicester_data %>% 
  select(cohort_number, birth_year_cat, sex, ethnic, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, birth_order, birth_order_cat, breast_feeding, breast_feeding_duration, age_mother_education, age_father_education, age_parental_education, parental_education_cat, townsend_2001, townsend_2001_quintile, urban_rural, physical_activity_recommended_under_5, physical_activity_recommended_5_to_18, mean_physical_activity_hrs_per_week_5_to_18, physical_activity_cat_mean, mean_fresh_fruit_per_week, fresh_fruit_cat, smoking_others_ever, smoking_personal_ever) %>% 
  tbl_summary(type = c(all_dichotomous() ~ "categorical",
              all_continuous() ~ "continuous2"),
              statistic = all_continuous() ~ "{mean} ({sd})",
              missing = "always",
              missing_stat = "{N_miss} ({p_miss}%)")

```

Comparing included vs. excluded participants.

```{r}
#| warning: false

Leicester_data |> 
  select(inclusion, cohort_number, birth_year_cat, sex, ethnic, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, birth_order, birth_order_cat, breast_feeding, breast_feeding_duration, age_mother_education, age_father_education, age_parental_education, parental_education_cat, townsend_2001, townsend_2001_quintile, urban_rural, physical_activity_recommended_under_5, physical_activity_recommended_5_to_18, mean_physical_activity_hrs_per_week_5_to_18, physical_activity_cat_mean, mean_fresh_fruit_per_week, fresh_fruit_cat, smoking_others_ever, smoking_personal_ever) |> 
  tbl_summary(by = inclusion,
              type = c(all_dichotomous() ~ "categorical",
              all_continuous() ~ "continuous2"),
              statistic = all_continuous() ~ "{mean} ({sd})",
              missing_stat = "{N_miss} ({p_miss}%)") |> 
  add_overall()

```

### Correlations of covariates

```{r}

Leicester_data |> 
  select(cohort_number, class, birth_year_cat, sex, ethnicity, ethnic, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, birth_order, breast_feeding, age_mother_education, age_father_education, age_parental_education, townsend_2001, townsend_2001_quintile, urbanization_2001, urban_rural, smoking_others_ever, mean_physical_activity_hrs_per_week_5_to_18, mean_fresh_fruit_per_week) |> 
  plot_correlation()

```

Ethnicity and age of mother at birth, education, sex, SES, breastfeeding, physical activity, fruits

```{r}

Leicester_data |> 
  select(ethnic, birth_year_cat, sex, townsend_2001, townsend_2001_quintile, age_parental_education, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, breast_feeding, mean_physical_activity_hrs_per_week_5_to_18, mean_fresh_fruit_per_week) |> 
  tbl_summary(by = "ethnic") |> 
  add_overall() |> 
  add_p()


```


# Check assumptions for CRA, IPW, and MI

## Directed acyclic graph

Draw DAG

```{r}

mdag_string <- 'dag {
M_age_mother_delivery [pos="-1.886,0.232"]
M_birth_order_cat [pos="-1.864,0.586"]
M_birth_weight [pos="-1.249,1.422"]
M_breast_feeding_duration [pos="-1.054,1.745"]
M_class [pos="0.817,-1.281"]
M_fresh_fruit_cat [pos="0.686,1.781"]
M_gestational_age [pos="-1.479,1.214"]
M_parental_education_cat [pos="-2.240,-0.284"]
M_physical_activity_cat [pos="0.027,1.794"]
M_smoking_mother_pregnancy [pos="-1.711,0.882"]
M_smoking_others_ever [pos="-0.462,1.816"]
age_mother_delivery [pos="-1.590,-0.003"]
birth_order_cat [pos="-1.459,0.289"]
birth_weight [pos="-0.993,1.250"]
birth_year_cat [pos="-1.800,-0.206"]
breast_feeding_duration [pos="-0.770,1.533"]
class [outcome,pos="1.030,0.121"]
ethnicity_cohort [exposure,pos="-2.232,-1.518"]
fresh_fruit_cat [pos="0.943,1.440"]
gestational_age [pos="-1.163,0.984"]
parental_education_cat [pos="-1.990,-0.622"]
physical_activity_cat [pos="0.020,1.382"]
sex [pos="-1.669,-1.689"]
smoking_mother_pregnancy [pos="-1.365,0.705"]
smoking_others_ever [pos="-0.474,1.431"]
townsend_2001 [pos="-2.052,-0.861"]
urban_rural [pos="-2.133,-1.096"]
age_mother_delivery -> birth_order_cat
age_mother_delivery -> birth_weight
age_mother_delivery -> breast_feeding_duration
age_mother_delivery -> class
age_mother_delivery -> fresh_fruit_cat
age_mother_delivery -> gestational_age
age_mother_delivery -> physical_activity_cat
age_mother_delivery -> smoking_others_ever
birth_order_cat -> birth_weight
birth_order_cat -> breast_feeding_duration
birth_order_cat -> class
birth_order_cat -> fresh_fruit_cat
birth_order_cat -> gestational_age
birth_order_cat -> physical_activity_cat
birth_order_cat -> smoking_others_ever
birth_weight -> breast_feeding_duration
birth_weight -> class
birth_weight -> fresh_fruit_cat
birth_weight -> physical_activity_cat
birth_weight -> smoking_others_ever
birth_year_cat -> M_class
birth_year_cat -> age_mother_delivery
birth_year_cat -> birth_order_cat
birth_year_cat -> birth_weight
birth_year_cat -> breast_feeding_duration
birth_year_cat -> class
birth_year_cat -> fresh_fruit_cat
birth_year_cat -> gestational_age
birth_year_cat -> physical_activity_cat
birth_year_cat -> smoking_mother_pregnancy
birth_year_cat -> smoking_others_ever
breast_feeding_duration -> M_breast_feeding_duration
breast_feeding_duration -> class
breast_feeding_duration -> fresh_fruit_cat
breast_feeding_duration -> physical_activity_cat
breast_feeding_duration -> smoking_others_ever
class -> M_class
ethnicity_cohort -> M_age_mother_delivery
ethnicity_cohort -> M_birth_order_cat
ethnicity_cohort -> M_birth_weight
ethnicity_cohort -> M_breast_feeding_duration
ethnicity_cohort -> M_class
ethnicity_cohort -> M_fresh_fruit_cat
ethnicity_cohort -> M_gestational_age
ethnicity_cohort -> M_parental_education_cat
ethnicity_cohort -> M_physical_activity_cat
ethnicity_cohort -> M_smoking_mother_pregnancy
ethnicity_cohort -> M_smoking_others_ever
ethnicity_cohort -> age_mother_delivery
ethnicity_cohort -> birth_weight
ethnicity_cohort -> birth_year_cat
ethnicity_cohort -> breast_feeding_duration
ethnicity_cohort -> class
ethnicity_cohort -> fresh_fruit_cat
ethnicity_cohort -> gestational_age
ethnicity_cohort -> parental_education_cat
ethnicity_cohort -> physical_activity_cat
ethnicity_cohort -> smoking_mother_pregnancy
ethnicity_cohort -> smoking_others_ever
ethnicity_cohort -> townsend_2001
ethnicity_cohort -> urban_rural
fresh_fruit_cat -> M_fresh_fruit_cat
fresh_fruit_cat -> class
gestational_age -> birth_weight
gestational_age -> breast_feeding_duration
gestational_age -> class
gestational_age -> fresh_fruit_cat
gestational_age -> physical_activity_cat
gestational_age -> smoking_others_ever
parental_education_cat -> M_age_mother_delivery
parental_education_cat -> M_birth_order_cat
parental_education_cat -> M_birth_weight
parental_education_cat -> M_breast_feeding_duration
parental_education_cat -> M_class
parental_education_cat -> M_fresh_fruit_cat
parental_education_cat -> M_gestational_age
parental_education_cat -> M_parental_education_cat
parental_education_cat -> M_physical_activity_cat
parental_education_cat -> M_smoking_mother_pregnancy
parental_education_cat -> M_smoking_others_ever
parental_education_cat -> age_mother_delivery
parental_education_cat -> birth_order_cat
parental_education_cat -> birth_weight
parental_education_cat -> breast_feeding_duration
parental_education_cat -> class
parental_education_cat -> fresh_fruit_cat
parental_education_cat -> gestational_age
parental_education_cat -> physical_activity_cat
parental_education_cat -> smoking_mother_pregnancy
parental_education_cat -> smoking_others_ever
physical_activity_cat -> M_physical_activity_cat
physical_activity_cat -> class
sex -> M_class
sex -> birth_weight
sex -> breast_feeding_duration
sex -> class
sex -> fresh_fruit_cat
sex -> gestational_age
sex -> physical_activity_cat
sex -> smoking_others_ever
smoking_mother_pregnancy -> M_smoking_mother_pregnancy
smoking_mother_pregnancy -> M_smoking_others_ever
smoking_mother_pregnancy -> birth_weight
smoking_mother_pregnancy -> class
smoking_mother_pregnancy -> fresh_fruit_cat
smoking_mother_pregnancy -> gestational_age
smoking_mother_pregnancy -> physical_activity_cat
smoking_mother_pregnancy -> smoking_others_ever
smoking_others_ever -> M_smoking_others_ever
smoking_others_ever -> class
townsend_2001 -> M_age_mother_delivery
townsend_2001 -> M_birth_order_cat
townsend_2001 -> M_birth_weight
townsend_2001 -> M_breast_feeding_duration
townsend_2001 -> M_class
townsend_2001 -> M_fresh_fruit_cat
townsend_2001 -> M_gestational_age
townsend_2001 -> M_parental_education_cat
townsend_2001 -> M_physical_activity_cat
townsend_2001 -> M_smoking_mother_pregnancy
townsend_2001 -> M_smoking_others_ever
townsend_2001 -> age_mother_delivery
townsend_2001 -> birth_order_cat
townsend_2001 -> birth_weight
townsend_2001 -> breast_feeding_duration
townsend_2001 -> class
townsend_2001 -> fresh_fruit_cat
townsend_2001 -> gestational_age
townsend_2001 -> parental_education_cat
townsend_2001 -> physical_activity_cat
townsend_2001 -> smoking_mother_pregnancy
townsend_2001 -> smoking_others_ever
urban_rural -> M_age_mother_delivery
urban_rural -> M_birth_order_cat
urban_rural -> M_birth_weight
urban_rural -> M_breast_feeding_duration
urban_rural -> M_class
urban_rural -> M_fresh_fruit_cat
urban_rural -> M_gestational_age
urban_rural -> M_parental_education_cat
urban_rural -> M_physical_activity_cat
urban_rural -> M_smoking_mother_pregnancy
urban_rural -> M_smoking_others_ever
urban_rural -> age_mother_delivery
urban_rural -> birth_order_cat
urban_rural -> birth_weight
urban_rural -> breast_feeding_duration
urban_rural -> class
urban_rural -> fresh_fruit_cat
urban_rural -> gestational_age
urban_rural -> parental_education_cat
urban_rural -> physical_activity_cat
urban_rural -> smoking_mother_pregnancy
urban_rural -> smoking_others_ever
urban_rural -> townsend_2001
}

'


dag_obj <- dagitty(mdag_string)

plot(dag_obj)



```

Check the missingness patterns

```{r}

md.pattern(Leicester_data |> 
  select(ethnicity_cohort, urban_rural, townsend_2001, parental_education_cat, birth_year_cat, sex, birth_order_cat, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, smoking_others_ever, physical_activity_cat, fresh_fruit_cat, class))
  

```


Check DAG with Leicester data

```{r}

Leicester_data_missing <- Leicester_data |> 
  mutate(M_parental_education_cat = factor(case_when(is.na(parental_education_cat) ~ 0,
                                              TRUE ~ 1)),
         M_birth_order_cat = case_when(is.na(birth_order_cat) ~ 0,
                                              TRUE ~ 1),
         M_smoking_mother_pregnancy = case_when(is.na(smoking_mother_pregnancy) ~ 0,
                                                TRUE ~ 1),
         M_age_mother_delivery = factor(case_when(is.na(age_mother_delivery) ~ 0,
                                                TRUE ~ 1)),
         M_gestational_age = case_when(is.na(gestational_age) ~ 0,
                                                TRUE ~ 1),
         M_birth_weight = case_when(is.na(birth_weight) ~ 0,
                                                TRUE ~ 1),
         M_breast_feeding_duration = case_when(is.na(breast_feeding_duration) ~ 0,
                                                TRUE ~ 1),
         M_smoking_others_ever = case_when(is.na(smoking_others_ever) ~ 0,
                                                TRUE ~ 1),
         M_physical_activity_cat = case_when(is.na(physical_activity_cat) ~ 0,
                                                TRUE ~ 1),
         M_fresh_fruit_cat = case_when(is.na(fresh_fruit_cat) ~ 0,
                                                TRUE ~ 1),
         M_class = case_when(is.na(class) ~ 0,
                                                TRUE ~ 1)) |> 
  select(ethnicity_cohort, urban_rural, townsend_2001, parental_education_cat, birth_year_cat, sex, birth_order_cat, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, smoking_others_ever, physical_activity_cat, fresh_fruit_cat, class,
         M_parental_education_cat, M_birth_order_cat, M_smoking_mother_pregnancy, M_age_mother_delivery, M_gestational_age, M_birth_weight, M_breast_feeding_duration, M_smoking_others_ever, M_physical_activity_cat, M_fresh_fruit_cat, M_class) |> 
  as.data.frame()

midoc::exploreDAG(mdag=mdag_string, data=Leicester_data_missing)



```


--> Very few of the tested independencies are actually dependent. This is probably due to the fact that in birth_year_cat 1985-1889 all participants are white.

Check basic analyses with midoc

```{r}

mdag_string <- 'dag {
class -> M_class
ethnicity -> M_class
ethnicity -> class
ethnicity -> townsend_2001
townsend_2001 -> M_class
townsend_2001 -> class
}'

midoc::checkCRA(y="class", covs="townsend_2001", r_cra="M_class", mdag=mdag_string)

midoc::checkMI(dep="class", preds="ethnicity townsend_2001", r_dep="M_class", mdag=mdag_string)

```

--> CRA is invalid because we condition on the missing outcome (complete records analysis) and the exposure (townsend_2001) affects this aswell. There is a backwards path open.
--> MI is invalid because the missing outcome is MNAR, because the outcome itself affects it.

Check analyses after IPW for complete sociodemographic factors with midoc

```{r}

mdag_string <- 'dag {
class -> M_class
ethnicity -> class
ethnicity -> townsend_2001
townsend_2001 -> class
}'

midoc::checkCRA(y="class", covs="townsend_2001", r_cra="M_class", mdag=mdag_string)

midoc::checkMI(dep="class", preds="ethnicity townsend_2001", r_dep="M_class", mdag=mdag_string)

```

--> The causal OR under CRA is asymptotically unbiased because only the outcome is missing and we closed the backwards path by IPW. The beta0 is biased however (prevalence of outcome).
--> MI is biased because the missing outcome is affected by the outcome itself.

--> The causal OR under CRA is also asymptotically unbiased for exposures of interest and confounders with missingness, as long as the exposure of interest (with missingness) does not affect the missingness of the outcome.
--> We limited this selection bias (missingness of the exposure --> missingness of the outcome) by IPW, but there is a possibility of residual selection bias, e.g. through parental education.


We will use IPW based on sociodemographic factors to minimize the selection bias.
CRA will be asymptotically unbiased for the case where missingness in the exposure is not linked to missingness in the outcome.
We will not use MI because the missingness in the outcome is most probably MNAR and because several exposures have missingness that is probably also MNAR.

# Weighting

## Design of LRC sample

It would be interesting to be representative of the pediatric and adolescent population in the UK. Unfortunately, we did only sample whites in 1990 and whites" and "south asians in 1998. Because we did not sample other ethnicities we cannot use design weights to be representative of the UK population of children. As described above for the missing outcome, this may affect the measures of the outcome (prevalence of BMI trajectory), but not the exposure ORs.

## Inclusion in BMI analysis

Create inverse probability weights for the included participants, to ensure the included children to the original Leicester cohort population.

### eFigure 1: DAGs before and after IPW, and possible residual biases

Selection bias would come from Sociodemographic factors. We use IPW to remove the connections from Sociodemographic factors to the missingness of the outcome.

Because we have missingness in parental education, but think that parental education could predict the missingness in the outcome, we will use an "unkown" category to also weight for parental education.

### eTable 3: Included vs. excluded

```{r}

Leicester_data |> select(inclusion, sex, birth_year_cat, ethnicity_cohort, parental_education_cat, urban_rural, townsend_2001, townsend_2001_quintile, birth_order_cat, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, breast_feeding_duration, smoking_others_ever, physical_activity_cat_mean, fresh_fruit_cat_mean, vegetables_cat_mean, crisps_cat_mean, chocolate_cat_mean) |> 
  mutate(inclusion = factor(inclusion, levels = c("included",  "excluded"))) |> 
  tbl_summary(by = "inclusion",
              missing = "always",
                 type = list(all_dichotomous() ~ "categorical",
                             all_continuous() ~ "continuous2"),
                 statistic = list(all_categorical() ~ "{n} ({p})",
                                  "townsend_2001" ~ "{median} ({p25}, {p75})",
                               "age_mother_delivery" ~ "{mean} ({sd})",
                               "gestational_age" ~ "{mean} ({sd})",
                               "birth_weight" ~ "{mean} ({sd})"),
                 digits = list(all_categorical() ~ 0,
                               all_continuous2() ~ 1),
                 missing_stat = "{N_miss} ({p_miss})") |> 
  add_overall() |> 
  add_p(test = list("townsend_2001" ~ "wilcox.test",
                    "age_mother_delivery" ~ "t.test",
                    "gestational_age" ~ "t.test",
                    "birth_weight" ~ "t.test")) |> 
  modify_fmt_fun(
    p.value = function(x) {
      ifelse(
        x < 0.0001, 
        "<0.0001", 
        format(round(x, 4), nsmall = 4))  # Show 4 decimal places
    }
  )


```





### eTable 4: IPW model

```{r}

Leicester_data <- Leicester_data |> 
  mutate(parental_education_cat = case_when(is.na(parental_education_cat) ~ "unkown",
                                            TRUE ~ parental_education_cat))

Leicester_data <- Leicester_data |> 
  mutate(birth_year_cat = factor(birth_year_cat,
                                 levels = c("1993-1997", "1985-1989")))

```

```{r}
#| warning: false

incl_model <- glm(inclusion ~ ethnicity_cohort + birth_year_cat + sex + parental_education_cat + urban_rural + townsend_2001_quintile, family = binomial, data = Leicester_data)
summary(incl_model)

cbind(exp(summary(incl_model)$coef[,1]), exp(confint(incl_model)), exp(summary(incl_model)$coef[,2]))


```


```{r}

Leicester_data_weighted <- Leicester_data |> 
  filter(is.na(sex)==FALSE) |> 
  mutate(p_included = predict(incl_model, type = "response")) |> 
  mutate(p_marginal = mean(inclusion=="included")) |> 
  mutate(ipweight_stabilized = case_when(inclusion == "included" ~ p_marginal/p_included,
                                         inclusion == "excluded" ~ (1 - p_marginal) / (1 - p_included)))

```


Check the created weights

```{r}

Leicester_data_weighted |> 
  select(ipweight_stabilized) |> 
  tbl_summary(statistic = all_continuous() ~ "{mean} ({sd}); {median} ({p25}, {p75}) [{min}, {max}]")


```

```{r}

Leicester_data <- Leicester_data |> 
  mutate(parental_education_cat = case_when(parental_education_cat=="unkown" ~ NA,
                                            TRUE ~ parental_education_cat))


Leicester_data <- Leicester_data |> 
  mutate(birth_year_cat = factor(birth_year_cat,
                                 levels = c("1985-1989", "1993-1997")))

Leicester_data_weighted <- Leicester_data_weighted |> 
  mutate(parental_education_cat = case_when(parental_education_cat=="unkown" ~ NA,
                                            TRUE ~ parental_education_cat))


Leicester_data_weighted <- Leicester_data_weighted |> 
  mutate(birth_year_cat = factor(birth_year_cat,
                                 levels = c("1985-1989", "1993-1997")))

```

### Alternative Table: Included vs. excluded after weighting

```{r}
svydesign(id = ~1, data = Leicester_data_weighted, weights = ~ipweight_stabilized, strata = ~inclusion) |> 
  as_survey() |> 
  select(inclusion, sex, birth_year_cat, ethnicity_cohort, parental_education_cat, urban_rural, townsend_2001, townsend_2001_quintile, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, birth_order_cat, breast_feeding_duration, smoking_others_ever, physical_activity_cat_mean, fresh_fruit_cat_mean, vegetables_cat_mean, crisps_cat_mean, chocolate_cat_mean) |> 
  tbl_svysummary(missing = "always",
                 by = "inclusion",
                 type = list(all_dichotomous() ~ "categorical",
                             all_continuous() ~ "continuous2"),
                 statistic = list(all_categorical() ~ "{n_unweighted} ({p})",
                                  "townsend_2001" ~ "{median} [{p25}, {p75}]",
                               "age_mother_delivery" ~ "{mean} ({sd})",
                               "gestational_age" ~ "{mean} ({sd})",
                               "birth_weight" ~ "{mean} ({sd})"),
                 digits = list(everything() ~ 0),
                 missing_stat = "{N_miss} ({p_miss})") |> 
  add_ci(pattern = "{stat} [{ci}]",
    style_fun = list(everything() ~ label_style_sigfig(digits = 5))) |> 
  add_p(test = list("townsend_2001" ~ "svy.wilcox.test",
                    "age_mother_delivery" ~ "svy.t.test",
                    "gestational_age" ~ "svy.t.test",
                    "birth_weight" ~ "svy.t.test")) |> 
  modify_fmt_fun(
    p.value = function(x) {
      ifelse(
        x < 0.0001, 
        "<0.0001", 
        format(round(x, 4), nsmall = 4))  # Show 4 decimal places
    }
  ) |> 
  add_overall()
```

### eFigure 2: Love plot of standardized differences


```{r}


covs <- Leicester_data_weighted[, c("sex", "birth_year_cat", "ethnicity_cohort", 
                                    "parental_education_cat", "urban_rural", 
                                    "townsend_2001", "townsend_2001_quintile", 
                                    "smoking_mother_pregnancy", "age_mother_delivery", 
                                    "gestational_age", "birth_weight", "birth_order_cat", 
                                    "breast_feeding_duration", "smoking_others_ever", 
                                    "physical_activity_cat_mean", "fresh_fruit_cat_mean", "vegetables_cat_mean", "crisps_cat_mean", "chocolate_cat_mean")]


# 2. Run love.plot manually mapping the pieces
love_plot_data <- love.plot(
  x = covs,
  treat = Leicester_data_weighted$inclusion,
  weights = Leicester_data_weighted$ipweight_stabilized,
  estimand = "ATE",
  stats = "mean.diffs",
  drop.missing = TRUE,
  binary = "std",
  continuous = "std",
  abs = TRUE,
  thresholds = c(m = .1),
  var.order = "unadjusted",
  line = TRUE,
  sample.names = c("Before IPW", "After IPW")
)

love_plot_data <- love_plot_data$data

love_plot_data <- love_plot_data |> 
  filter(!grepl("NA", var)) |> 
  filter(var != "townsend_2001")

love_plot_data <- love_plot_data |> 
  bind_rows(love_plot_data |>
              filter(var == "sex_female" | var == "birth_year_cat_1993-1997" | var == "ethnicity_cohort_South Asian" | var == "urban_rural_rural" | var == "smoking_mother_pregnancy_yes" | var == "smoking_others_ever_yes" | var == "fresh_fruit_cat_mean_less than once daily" |
                       var == "vegetables_cat_mean_less than once daily" | var == "crisps_cat_mean_less than once daily" | var == "chocolate_cat_mean_less than once daily")
            |> mutate(var = case_when(var == "sex_female" ~ "sex_male",
                                      var == "birth_year_cat_1993-1997" ~ "birth_year_cat_1985-1989",
                                      var == "ethnicity_cohort_South Asian" ~ "ethnicity_cohort_White",
                                      var == "urban_rural_rural" ~ "urban_rural_urban",
                                      var == "smoking_mother_pregnancy_yes" ~ "smoking_mother_pregnancy_no",
                                      var == "smoking_others_ever_yes" ~ "smoking_others_ever_no",
                                      var == "fresh_fruit_cat_mean_less than once daily" ~ "'Once daily or more'",
                                      var == "vegetables_cat_mean_less than once daily" ~ "' Once daily or more'",
                                      var == "crisps_cat_mean_less than once daily" ~ "'  Once daily or more'",
                                      var == "chocolate_cat_mean_less than once daily" ~ "'   Once daily or more'")))

love_plot_data <- love_plot_data |> 
  mutate(var = case_when(var == "sex_female" ~ "Female",
                           var == "sex_male" ~ "Male",
                          var == "birth_year_cat_1985-1989" ~ "'1985–1989'",
                          var == "birth_year_cat_1993-1997" ~ "'1993–1997'",
                          var == "ethnicity_cohort_South Asian" ~ "'South Asian'",
                          var == "ethnicity_cohort_White" ~ "White",
                          var == "urban_rural_rural" ~ "'Suburbs and rural'",
                          var == "urban_rural_urban" ~ "'Inner city'",
                          var == "townsend_2001_quintile_1" ~ "'Quintile 1'",
                          var == "townsend_2001_quintile_2" ~ "'Quintile 2'",
                          var == "townsend_2001_quintile_3" ~ "'Quintile 3'",
                          var == "townsend_2001_quintile_4" ~ "'Quintile 4'",
                          var == "townsend_2001_quintile_5" ~ "'Quintile 5'",
                          var == "parental_education_cat_compulsory education" ~ "Compulsory",
                          var == "parental_education_cat_further education" ~ "Further",
                          var == "parental_education_cat_higher education" ~ "Higher",
                          var == "smoking_mother_pregnancy_yes" ~ "Yes",
                          var == "smoking_mother_pregnancy_no" ~ "No",
                          var == "age_mother_delivery" ~ "bold('Age of mother at birth')",
                          var == "gestational_age" ~ "bold('Gestational age')",
                          var == "birth_weight" ~ "bold('Birth weight')",
                          var == "birth_order_cat_first child" ~ "'First child'",
                          var == "birth_order_cat_second child" ~ "'Second child'",
                          var == "birth_order_cat_third or later child" ~ "'Third or later child'",
                          var == "breast_feeding_duration_0 months" ~ "'0 months'",
                          var == "breast_feeding_duration_<=1 month" ~ "'≤1 month'",
                          var == "breast_feeding_duration_1-3 months" ~ "'1–3 months'",
                          var == "breast_feeding_duration_4-6 months" ~ "'4–6 months'",
                          var == "breast_feeding_duration_>=6 months" ~ "'≥6 months'",
                          var == "physical_activity_cat_mean_0-1 hr per week" ~ "'0–1 hr/week'",
                          var == "physical_activity_cat_mean_2-3 hrs per week" ~ "'2–3 hrs/week'",
                          var == "physical_activity_cat_mean_4-5 hrs per week" ~ "'4–5 hrs/week'",
                          var == "physical_activity_cat_mean_>=6 hrs per week" ~ "'≥6 hrs/week'",
                          var == "fresh_fruit_cat_mean_less than once daily" ~ "'Less than once daily'",
                          var == "vegetables_cat_mean_less than once daily" ~ "' Less than once daily'",
                          var == "crisps_cat_mean_less than once daily" ~ "'  Less than once daily'",
                          var == "chocolate_cat_mean_less than once daily" ~ "'   Less than once daily'",
                          var == "smoking_others_ever_yes" ~ "' Yes'",
                          var == "smoking_others_ever_no" ~ "' No'",
                          TRUE ~ var),
         transparency = "filled")


love_plot_variable_titles <- data.frame(var = c("bold(Sex)",
                                                  "bold('Birth year')",
                                                  "bold(Ethnicity)",
                                                  "bold('Degree of urbanization')",
                                                  "bold('Townsend index')",
                                                  "bold('Parental education')",
                                                  
                                                  "bold('Birth order')",
                                                  "bold('Smoking during pregnancy')",
                                                  "bold(Breastfeeding)",
                                                  
                                                  "bold('Secondhand tobacco smoke')",
                                                  "bold('Physical activity')",
                                                  "bold('Fresh fruit intake')",
                                                  "bold('Vegetable intake')",
                                                  "bold('Crisps intake')",
                                                  "bold('Chocolates intake')"),
                                         stat = 0,
                                        transparency = "transparent")


love_plot_data <- love_plot_data |> 
  bind_rows(love_plot_variable_titles)

love_plot_data <- love_plot_data |> 
  mutate(label = factor(var,
                      levels = c(
                        "bold(Sex)",
                        "Male",
                        "Female",
                        "bold('Birth year')",
                        "'1985–1989'",
                        "'1993–1997'",
                        "bold(Ethnicity)",
                        "White",
                        "'South Asian'",
                        "bold('Degree of urbanization')",
                        "'Inner city'",
                        "'Suburbs and rural'",
                        "bold('Townsend index')",
                        "'Quintile 1'",
                        "'Quintile 2'",
                        "'Quintile 3'",
                        "'Quintile 4'",
                        "'Quintile 5'",
                        "bold('Parental education')",
                        "Compulsory",
                        "Further",
                        "Higher",
                        "bold('Birth order')",
                        "'First child'",
                        "'Second child'",
                        "'Third or later child'",
                        "bold('Age of mother at birth')",
                        "bold('Gestational age')",
                        "bold('Birth weight')",
                        "bold('Smoking during pregnancy')",
                        "No",
                        "Yes",
                        "bold(Breastfeeding)",
                        "'0 months'",
                        "'≤1 month'",
                        "'1–3 months'",
                        "'4–6 months'",
                        "'≥6 months'",
                        "bold('Secondhand tobacco smoke')",
                        "' No'",
                        "' Yes'",
                        "bold('Physical activity')",
                        "'0–1 hr/week'",
                        "'2–3 hrs/week'",
                        "'4–5 hrs/week'",
                        "'≥6 hrs/week'",
                        "bold('Fresh fruit intake')",
                        "'Once daily or more'",
                        "'Less than once daily'",
                        "bold('Vegetable intake')",
                        "' Once daily or more'",
                        "' Less than once daily'",
                        "bold('Crisps intake')",
                        "'  Once daily or more'",
                        "'  Less than once daily'",
                        "bold('Chocolates intake')",
                        "'   Once daily or more'",
                        "'   Less than once daily'")),
         label = fct_rev(label))


love_plot <- love_plot_data |> ggplot(aes(x = stat, y = label, colour = Sample))+
  geom_point(aes(alpha = transparency))+
  geom_vline(xintercept = 0.1, linewidth = 1, color = "black", linetype = "dashed")+
  scale_x_continuous(limits = c(0, 0.3), breaks = c(0, 0.1, 0.2, 0.3))+
  scale_y_discrete(labels = function(x) parse(text = x))+
  scale_colour_manual(breaks = c("Before IPW", "After IPW"), values = c(unibeRed(), unibeIce(), NA))+
  scale_alpha_manual(breaks = c("transparent", "filled"), values = c(0, 100), guide = "none")+
  xlab("Standardized difference")+
  theme_minimal()+
  theme(panel.grid.major.y = element_blank(),
        legend.title = element_blank(),
        legend.position = "top",
        legend.text = element_text(size = 10, face = "bold", color = "black"),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 10, face = "bold", color = "black"),
        axis.text.y = element_text(size = 8, color = "black"),
        axis.text.x = element_text(size = 8, color = "black"))



ggsave(love_plot, filename = here("output", "figures", "love_plot.svg"), device = "svg", units = "cm", bg = "white", width = 12, height = 20)  
  

```



# Main analysis

## Table 1: Study characteristics

```{r}

Leicester_data |> 
  select(inclusion, sex, birth_year_cat, ethnicity_cohort, parental_education_cat, urban_rural, townsend_2001, townsend_2001_quintile, birth_order_cat, age_mother_delivery, smoking_mother_pregnancy, gestational_age, birth_weight, breast_feeding, smoking_others_ever, physical_activity_cat_minimal, fresh_fruit_cat_mean, vegetables_cat_mean, crisps_cat_mean, chocolate_cat_mean) |> 
  filter(inclusion == "included") |>
  select(-inclusion) |> 
  tbl_summary(missing = "always",
              by = "ethnicity_cohort",
                 type = list(all_dichotomous() ~ "categorical",
                             all_continuous() ~ "continuous2"),
                 statistic = list(all_categorical() ~ "{n} ({p})",
                                  "townsend_2001" ~ "{median} ({p25}, {p75})",
                               "age_mother_delivery" ~ "{mean} ({sd})",
                               "gestational_age" ~ "{mean} ({sd})",
                               "birth_weight" ~ "{mean} ({sd})"),
                 digits = list(all_categorical() ~ 0,
                               all_continuous2() ~ 1),
                 missing_stat = "{N_miss} ({p_miss})") |> 
  add_overall() |> 
  add_p(test = list("townsend_2001" ~ "wilcox.test",
                    "age_mother_delivery" ~ "t.test",
                    "gestational_age" ~ "t.test",
                    "birth_weight" ~ "t.test")) |> 
  modify_fmt_fun(
    p.value = function(x) {
      ifelse(
        x < 0.0001, 
        "<0.0001", 
        format(round(x, 4), nsmall = 4))  # Show 4 decimal places
    }
  )



```

### Alternative Table: By cohort number

```{r}

Leicester_data |> 
  select(inclusion, cohort_number, sex, ethnicity_cohort, parental_education_cat, urban_rural, townsend_2001, townsend_2001_quintile, smoking_mother_pregnancy, age_mother_delivery, gestational_age, term_birth, birth_weight, birth_order_cat, breast_feeding_duration, smoking_others_ever, physical_activity_cat_mean, fresh_fruit_cat_mean) |> 
  tbl_summary(by = cohort_number,
              type = c(all_dichotomous() ~ "categorical",
                       all_continuous() ~ "continuous2"),
              statistic = list(all_continuous() ~ "{median} ({p25}, {p75})",
                               all_categorical() ~ "{n} ({p})"),
              missing_stat = "{N_miss} ({p_miss}%)",
              digits = everything() ~ 0)

```

## Present BMI Trajectories

### eSupplement-Table 7: Summary of model fit

Write function to compare summary statistics of models.

```{r}
#| warning: false

summary_models <- function(models){
  
  summary_df <- data.frame(
    Model = character(),
    G = integer(),
    npm = integer(),
    loglik = numeric(),
    BIC = numeric(),
    SABIC = numeric(),
    entropy = numeric(),
    class_percentages = character(),
    avg_post_prob = character(),
    prop_above_post_prob = character(),
    stringsAsFactors = FALSE
  )
  
  for (i in seq_along(models)) {
    model <- models[[i]]
    
    model_name <- names(models)[i]
    
    capture.output(suppressMessages(suppressWarnings(summary_stats <- summarytable(model, which = c("G", "loglik", "npm", "BIC", "SABIC", "entropy", "%class")))))
    
    class_cols <- grep("^%class", colnames(summary_stats), value = TRUE)
    
    class_percentages <- paste(round(summary_stats[, class_cols], 2), collapse = ", ")
    
    avg_post_prob <- c()
    
    if (summary_stats[,"G"] > 1){
      for (n_class in 1:summary_stats[,"G"]){
      
      capture.output(suppressMessages(suppressWarnings(avg_post_prob <- c(avg_post_prob,postprob(model, threshold = 0.7)[[2]][n_class, n_class]))))
      
      }
      
      capture.output(suppressMessages(suppressWarnings(prop_above_post_prob <- postprob(model, threshold = 0.7)[[3]]/100)))
    
    } else {
      avg_post_prob = 1
      prop_above_post_prob = 1
    }
    
    avg_post_prob <- paste(as.character(avg_post_prob), collapse = ", ")
    prop_above_post_prob <- paste(as.character(prop_above_post_prob), collapse = ", ")
    
    summary_df <- rbind(summary_df, data.frame(
      Model = model_name,
      G = summary_stats[,"G"],
      npm = summary_stats[,"npm"],
      loglik = summary_stats[,"loglik"],
      BIC = summary_stats[,"BIC"],
      SABIC = summary_stats[,"SABIC"],
      entropy = summary_stats[,"entropy"],
      class_percentages = class_percentages,
      avg_post_prob = avg_post_prob,
      prop_above_post_prob = prop_above_post_prob,
      stringsAsFactors = FALSE
    ))
  }
  
  return(summary_df)
}

two_knot_models_total <- c(bmi_models_total[1], bmi_models_total[3], bmi_models_total[5], bmi_models_total[7], bmi_models_total[9], bmi_models_total[11], bmi_models_total[13])

summary_models(two_knot_models_total)

```


### Alternative Table: Identified trajectories and number of participants

```{r}

Leicester_data |> select(class, ethnicity_cohort) |> 
  tbl_summary(by = "ethnicity_cohort") |> 
  add_overall()

```


### Plot BMI trajectories

Functions

```{r}
#| warning: false

plot_hlme_model_bmi <- function(hlme_model, age = 18){

num_classes <- hlme_model$ng
num_subjects <- hlme_model$ns
  
data_predict <- data.frame(age_days = seq(14, trunc(age*365.25))) %>% 
  mutate(age_years = age_days/365.25)

predictions <- predictY(hlme_model, data_predict, var.time = "age_years", draws = TRUE)
  
data_predict <- data.frame(predictions[2], predictions[1])
  
if (num_classes == 1){
  data_predict <- data_predict %>% 
    rename("pred.Ypred_class1" = "pred.Ypred", "pred.lower.Ypred_class1" = "pred.lower.Ypred", "pred.upper.Ypred_class1" = "pred.upper.Ypred")
  
}

rename_columns <- setNames(
  paste0("pred.Ypred_class", 1:num_classes), 
  paste0("mean_class", 1:num_classes)
)
  
col_names <- names(data_predict)
  
for (i in 1:num_classes) {
    # Adjust column names based on the pattern in your `predictions`
    mean_col <- paste0("pred.Ypred_class", i)
    if (mean_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == mean_col)] <- paste0("mean_class", i)
    }
  }

for (i in 1:num_classes) {
    lower_col <- paste0("pred.lower.Ypred_class", i)
    upper_col <- paste0("pred.upper.Ypred_class", i)
    
    if (lower_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == lower_col)] <- paste0("lower_class", i)
    }
    
    if (upper_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == upper_col)] <- paste0("upper_class", i)
    }
  }

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

title <- paste0(num_classes, "-class model")

if (num_classes == 1){
  color_values = unibeSaphire()
}

if (num_classes == 2){
  color_values = c(unibeSaphire(), unibeApricot())
}

if (num_classes == 3){
  color_values = c(unibeSaphire(), unibeIce(), unibeRed())
}

if (num_classes == 4){
  color_values = c(unibeRed(), unibeIce(), unibeSaphire(), unibeApricot())
}

if (num_classes == 5){
  color_values = c(unibeMustard(), unibeIce(), unibeApricot(), unibeSaphire(), unibeRed())
}

if (num_classes == 6){
  color_values = c(unibeMint(), unibeRed(), unibeSaphire(), unibeIce(), unibeApricot(), unibeMustard())
}

if (num_classes == 7){
  color_values = c(unibeRed(), "#E93A1F", unibeApricot(), unibeIce(), unibeMint(), unibeSaphire(), unibeMustard())
}

if (num_subjects == 2661){
  color_values = c(unibeRed(), unibeIce(), unibeSaphire(), unibeMustard(), unibeApricot())
}

if (num_subjects == 2910){
  color_values = c(unibeRed(), unibeIce(), unibeMustard(), unibeSaphire(), unibeApricot())
}

if (num_subjects == 3871){
  color_values = c(unibeRed(), unibeMustard(), unibeIce(), unibeSaphire(), unibeApricot())
}

if (num_subjects == 1700){
  color_values = c(unibeApricot(), unibeRed(), unibeSaphire(), unibeMustard(), unibeIce())
}


plot_bmi_trajectories <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.5, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  new_scale_color()+
  new_scale_fill()+
  geom_ribbon(data = data_predict, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_predict, aes(x = age_years, y = mean, color = class), linewidth = 0.5) +
  scale_fill_manual(name = "Trajectories",
                    values = color_values,
                    guide = guide_legend(order = 2)) +
  scale_color_manual(name = "Trajectories",
                    values = color_values,
                    guide = guide_legend(order = 2)) +
  labs(x = "Age (years)",
       y = (bquote(BMI~(kg/m^2))))+
  scale_x_continuous(limits = c(0, 18), expand = c(0,0), breaks = c(0,5,10,15,18))+
  scale_y_continuous(expand = c(0,0), breaks = c(10,15,20,25,30,35,40))+
  theme_minimal()+
  theme(axis.line = element_line(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "none",
        axis.title = element_text(family = "Arial", size = 8),
        axis.text = element_text(family = "Arial", size = 7),
        plot.margin = margin(l = 0, t = 0, r = 0.5, b = 1, unit = "cm"))

return(plot_bmi_trajectories)

}
```


Plot of predicted trajectories

```{r}
#| warning: false

best_model <- bmi_models_total[[9]]

plot_hlme_model_bmi(best_model)

```

#### Alternative Figure: BMI trajectories

Prepare BMI trajectory data

```{r}

data_predict <- data.frame(age_days = seq(0, trunc(18*365.25))) %>% 
  mutate(age_years = age_days/365.25)

predictions <- predictY(best_model, data_predict, var.time = "age_years", draws = TRUE)
  
data_predict <- data.frame(predictions[2], predictions[1])
  

rename_columns <- setNames(
  paste0("pred.Ypred_class", 1:5), 
  paste0("mean_class", 1:5)
)
  
col_names <- names(data_predict)
  
for (i in 1:5) {
    # Adjust column names based on the pattern in your `predictions`
    mean_col <- paste0("pred.Ypred_class", i)
    if (mean_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == mean_col)] <- paste0("mean_class", i)
    }
  }

for (i in 1:5) {
    lower_col <- paste0("pred.lower.Ypred_class", i)
    upper_col <- paste0("pred.upper.Ypred_class", i)
    
    if (lower_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == lower_col)] <- paste0("lower_class", i)
    }
    
    if (upper_col %in% col_names) {
      colnames(data_predict)[which(colnames(data_predict) == upper_col)] <- paste0("upper_class", i)
    }
  }

data_predict <- data_predict %>% 
  pivot_longer(cols = -age_years,
               names_to = c(".value", "class"),
               names_sep = "_",
               values_to = "value")

data_predict <- data_predict |> 
  mutate(class = factor(class, levels = c("class5", "class3", "class1", "class4", "class2"), labels = c("class5", "class3", "class1", "class4", "class2")))

```

Plot trajectories

```{r}

plot_bmi_trajectories <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.75, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  new_scale_color()+
  new_scale_fill()+
  geom_ribbon(data = data_predict, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_predict, aes(x = age_years, y = mean, color = class), linewidth = 0.75) +
  scale_fill_manual(name = "Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 2)) +
  scale_color_manual(name = "Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 2)) +
  labs(x = "Age (years)",
       y = (bquote(BMI~(kg/m^2))))+
  scale_x_continuous(limits = c(0, 18), expand = c(0,0), breaks = c(0,5,10,15,18))+
  scale_y_continuous(expand = c(0,0), breaks = c(15,20,25,30))+
  theme_minimal()+
  theme(axis.line = element_line(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "none",
        axis.title = element_text(family = "Arial", size = 11),
        axis.text = element_text(family = "Arial", size = 8),
        plot.margin = margin(l = 0, t = 0, r = 0.5, b = 0, unit = "cm"))

plot_bmi_trajectories

ggsave(plot_bmi_trajectories, filename = here("output", "figures", "plot_bmi_trajectories.svg"), device = "svg", units = "cm", height = 15, width = 16, bg = "white")

plot_legend <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.25, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  new_scale_color()+
  new_scale_fill()+
  geom_ribbon(data = data_predict, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_predict, aes(x = age_years, y = mean, color = class), linewidth = 0.5) +
  scale_fill_manual(name = "Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 2)) +
  scale_color_manual(name = "Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 2)) +
  labs(x = "Age (years)",
       y = (bquote(Mean~BMI~(kg/m^2))))+
  scale_x_continuous(limits = c(14/365.25, 18), expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))+
  theme_minimal()+
  theme(legend.title = element_text(family = "Helvetica", face = "bold", size = 8),
        legend.text = element_markdown(family = "Helvetica", size = 8),
        legend.margin = margin(l = 0, t = 0, unit = "cm"),
        legend.box.spacing = unit(0.5,"cm"),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.25, "cm"))

plot_legend <- get_legend(plot_legend) 

plot_legend

plot_bmi_trajectories_legend <- plot_grid(plot_bmi_trajectories, plot_legend, ncol=1, rel_heights = c(1, 0.5))

plot_bmi_trajectories_legend

ggsave(plot_bmi_trajectories_legend, filename = here("output", "figures", "plot_bmi_trajectories_legend.svg"), device = "svg", units = "cm", bg = "white", width = 16, height = 14)


```


#### Figure 2: Observed vs. predicted trajectories

Prepare the observed data

```{r}

age_means <- Leicester_anthro_18 |> 
  mutate(age_group = ntile(age_years, 10)) |> 
  group_by(age_group) |> 
  summarise(mean_age = mean(age_years, na.rm = TRUE), .groups = "drop")

bmi_summary <- Leicester_anthro_18 |> 
  left_join(Leicester_data |> select(serialno, class), by = "serialno") |> 
  mutate(age_group = ntile(age_years, 10)) |> 
  group_by(age_group, class) |> 
  summarise(
    mean_bmi = mean(bmi, na.rm = TRUE),
    n = n(),
    sd_bmi = sd(bmi, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    se_bmi = sd_bmi / sqrt(n),
    ci_lower = mean_bmi - qt(0.975, df = n - 1) * se_bmi,
    ci_upper = mean_bmi + qt(0.975, df = n - 1) * se_bmi
  )

data_observed <- bmi_summary |> 
  left_join(age_means, by = "age_group") |> 
  mutate(class = factor(class, levels = c("Childhood onset obesity", "Adolescent onset overweight", "Early overweight resolving", "Stable normal BMI", "Persistent low BMI")))

```


```{r}
#| warning: false

plot_observed_bmi_trajectories <- ggplot() +
    geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.75, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  new_scale_color()+
  new_scale_fill()+
  geom_point(data = data_observed, mapping = aes(x = mean_age, y = mean_bmi, color = class))+
  geom_errorbar(data = data_observed, mapping = aes(x = mean_age, ymin = ci_lower, ymax = ci_upper, color = class))+
  scale_color_manual(name = "Observed BMI",
                    values = c("Childhood onset obesity" = unibeRed(),
                               "Adolescent onset overweight" = unibeApricot(),
                               "Early overweight resolving" = unibeMustard(),
                               "Stable normal BMI" = unibeSaphire(),
                               "Persistent low BMI" = unibeIce()),
                    labels = c("Childhood onset obesity" = "Childhood onset obesity",
                               "Adolescent onset overweight" = "Adolescent onset overweight",
                               "Early overweight resolving" = "Early overweight resolving",
                               "Stalbe normal BMI" = "Stalbe normal BMI",
                               "Persistent low BMI" = "Persistent low BMI"),
                    guide = guide_legend(order = 2)) +
  new_scale_color()+
  new_scale_fill()+
  geom_ribbon(data = data_predict, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_predict, aes(x = age_years, y = mean, color = class), linewidth = 0.75) +
  scale_fill_manual(name = "Predicted BMI Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 3)) +
  scale_color_manual(name = "Predicted BMI Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 3)) +
  labs(x = "Age (years)",
       y = (bquote(BMI~(kg/m^2))))+
  scale_x_continuous(limits = c(0, 18), expand = c(0,0), breaks = c(0,5,10,15,18))+
  scale_y_continuous(limits = c(12, 31), expand = c(0,0), breaks = c(15,20,25,30))+
  theme_minimal()+
  theme(axis.line = element_line(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        legend.position = "none",
        axis.title = element_text(family = "Arial", size = 11),
        axis.text = element_text(family = "Arial", size = 8),
        plot.margin = margin(l = 0, t = 1, r = 0.5, b = 1, unit = "cm"))

plot_observed_bmi_trajectories

ggsave(plot_observed_bmi_trajectories, filename = here("output", "figures", "plot_observed_bmi_trajectories.svg"), device = "svg", units = "cm", height = 15, width = 16, bg = "white")

plot_legend_observed <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.25, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  new_scale_color()+
  new_scale_fill()+
  geom_point(data = data_observed, mapping = aes(x = mean_age, y = mean_bmi, color = class))+
  geom_errorbar(data = data_observed, mapping = aes(x = mean_age, ymin = ci_lower, ymax = ci_upper, color = class))+
  scale_color_manual(name = "Observed BMI",
                    values = c("Childhood onset obesity" = unibeRed(),
                               "Adolescent onset overweight" = unibeApricot(),
                               "Early overweight resolving" = unibeMustard(),
                               "Stable normal BMI" = unibeSaphire(),
                               "Persistent low BMI" = unibeIce()),
                    labels = c("Childhood onset obesity" = "Childhood onset obesity",
                               "Adolescent onset overweight" = "Adolescent onset overweight",
                               "Early overweight resolving" = "Early overweight resolving",
                               "Stable normal BMI" = "Stable normal BMI",
                               "Persistent low BMI" = "Persistent low BMI"),
                    guide = guide_legend(order = 2, nrow = 5)) +
  new_scale_color()+
  new_scale_fill()+
  geom_ribbon(data = data_predict, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_predict, aes(x = age_years, y = mean, color = class), linewidth = 0.5) +
  scale_fill_manual(name = "Predicted BMI Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 3, nrow = 5)) +
  scale_color_manual(name = "Predicted BMI Trajectories",
                    values = c("class5" = unibeRed(),
                               "class3" = unibeApricot(),
                               "class1" = unibeMustard(),
                               "class4" = unibeSaphire(),
                               "class2" = unibeIce()),
                    labels = c("class5" = "Childhood onset obesity",
                               "class3" = "Adolescent onset overweight",
                               "class1" = "Early overweight resolving",
                               "class4" = "Stable normal BMI",
                               "class2" = "Persistent low BMI"),
                    guide = guide_legend(order = 3, nrow = 5)) +
  labs(x = "Age (years)",
       y = (bquote(Mean~BMI~(kg/m^2))))+
  scale_x_continuous(limits = c(14/365.25, 18), expand = c(0,0), )+
  scale_y_continuous(expand = c(0,0))+
  theme_minimal()+
  theme(legend.title = element_text(family = "Helvetica", face = "bold", size = 8),
        legend.text = element_markdown(family = "Helvetica", size = 8),
        legend.margin = margin(l = 0, t = 0, unit = "cm"),
        legend.box.spacing = unit(0.5,"cm"),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.25, "cm"),
        legend.box = "horizontal")

plot_legend_observed <- get_legend(plot_legend_observed) 

plot_legend_observed

plot_observed_bmi_trajectories_legend <- plot_grid(plot_observed_bmi_trajectories, plot_legend_observed, ncol=1, rel_heights = c(1, 0.15))

plot_observed_bmi_trajectories_legend

ggsave(plot_observed_bmi_trajectories_legend, filename = here("output", "figures", "plot_observed_bmi_trajectories_legend.svg"), device = "svg", units = "cm", bg = "white", width = 16, height = 14)
  

```




#### eSupplement Figure 3: Female vs. Male / White vs. South Asian

```{r}
#| warning: false

plot_females <- plot_hlme_model_bmi(bmi_models_female[[9]])
plot_females

plot_males <- plot_hlme_model_bmi(bmi_models_male[[9]])
plot_males

plot_europe <- plot_hlme_model_bmi(bmi_models_europe[[9]])
plot_europe

plot_asia <- plot_hlme_model_bmi(bmi_models_asia[[9]])
plot_asia

plot_5_class <- plot_hlme_model_bmi(bmi_models_total[[9]])
plot_5_class

plot_legend_models <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.25, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  theme_minimal()+
  theme(legend.title = element_text(family = "Helvetica", face = "bold", size = 8),
        legend.text = element_markdown(family = "Helvetica", size = 8),
        legend.margin = margin(l = 0, t = 0, unit = "cm"),
        legend.box.spacing = unit(0.5,"cm"),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.25, "cm"))

legend_models <- get_legend(plot_legend_models) 

plot_female_male_european_asian <- plot_grid(plot_females, plot_males, plot_europe, plot_asia, plot_5_class, legend_models,
                         ncol = 2,
                         labels = c("Female ", "Male ", "White", "South Asian", "Combined", ""),
                         label_x = c(0.12, 0.12, 0.12, 0.12, 0.12),
                         hjust = c(0,0,0,0,0),
                         label_size = 12) + theme(plot.margin = margin(t = 0.5, unit = "cm"))

plot_female_male_european_asian

ggsave(plot_female_male_european_asian, filename = here("output", "figures", "plot_female_male_european_asian.svg"), device = "svg", units = "cm", bg = "white", width = 16, height = 18)

```



#### eSupplement Figure 4: All 7 models

```{r}
#| warning: false

plot_1_class <- plot_hlme_model_bmi(bmi_models_total[[1]])
plot_1_class

plot_2_class <- plot_hlme_model_bmi(bmi_models_total[[3]])
plot_2_class

plot_3_class <- plot_hlme_model_bmi(bmi_models_total[[5]])
plot_3_class

plot_4_class <- plot_hlme_model_bmi(bmi_models_total[[7]])
plot_4_class

plot_5_class <- plot_hlme_model_bmi(bmi_models_total[[9]])
plot_5_class

plot_6_class <- plot_hlme_model_bmi(bmi_models_total[[11]])
plot_6_class

plot_7_class <- plot_hlme_model_bmi(bmi_models_total[[13]])
plot_7_class

plot_legend_models <- ggplot() +
  geom_ribbon(data = data_bmi_reference, mapping = aes(x = age_years, ymin = lower, ymax = upper, fill = class), alpha = 0.3) +
  geom_line(data = data_bmi_reference, mapping = aes(x = age_years, y = mean, color = class), linewidth = 0.25, linetype = "dashed") +
  scale_fill_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "grey5",
                               "overweight" = "grey40",
                               "50th percentile" = "NA",
                               "underweight" = "grey80"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  scale_color_manual(name = "UK-WHO growth charts",
                    values = c("obesity" = "NA",
                               "overweight" = "NA",
                               "50th percentile" = "black",
                               "underweight" = "NA"),
                    labels = c("obesity" = "Obesity above 98<sup>th</sup> centile",
                                "overweight" = "Overweight above 91<sup>st</sup> centile",
                               "50th percentile" = "50<sup>th</sup> centile",
                               "underweight" = "Underweight below 2<sup>nd</sup> centile"),
                    guide = guide_legend(order = 1)) +
  theme_minimal()+
  theme(legend.title = element_text(family = "Helvetica", face = "bold", size = 8),
        legend.text = element_markdown(family = "Helvetica", size = 8),
        legend.margin = margin(l = 0, t = 0, unit = "cm"),
        legend.box.spacing = unit(0.5,"cm"),
        legend.key.width = unit(0.5, "cm"),
        legend.key.height = unit(0.25, "cm"))

legend_models <- get_legend(plot_legend_models) 

plot_models <- plot_grid(plot_1_class, plot_2_class, plot_3_class, plot_4_class, plot_5_class, plot_6_class, plot_7_class, legend_models,
                         ncol = 2,
                         labels = c(" 1 class ", "2 classes", "3 classes", "4 classes", "5 classes", "6 classes", "7 classes", ""),
                         label_size = 12)

plot_models

ggsave(plot_models, filename = here("output", "figures", "plot_models.svg"), device = "svg", units = "cm", bg = "white", width = 16, height = 23)

```




## Characterize BMI Trajectories

### Table 2: Participants' characteristics across trajectories

```{r}
#| warning: false

Leicester_data |> 
  select(class, sex, birth_year_cat, ethnicity_cohort, parental_education_cat, urban_rural, townsend_2001, townsend_2001_quintile, birth_order_cat, smoking_mother_pregnancy, age_mother_delivery, gestational_age, birth_weight, breast_feeding_duration, smoking_others_ever, physical_activity_cat_mean, fresh_fruit_cat_mean, vegetables_cat_mean, crisps_cat_mean, chocolate_cat_mean) |> 
  tbl_summary(by = class,
              missing = "always",
                 type = list(all_dichotomous() ~ "categorical",
                             all_continuous() ~ "continuous2"),
                 statistic = list(all_categorical() ~ "{n} ({p})",
                                  "townsend_2001" ~ "{median} ({p25}, {p75})",
                               "age_mother_delivery" ~ "{mean} ({sd})",
                               "gestational_age" ~ "{mean} ({sd})",
                               "birth_weight" ~ "{mean} ({sd})"),
                 digits = list(all_categorical() ~ 0),
                 missing_stat = "{N_miss} ({p_miss})") |> 
  add_p(test = list("townsend_2001" ~ "kruskal.test",
                    "age_mother_delivery" ~ "oneway.test",
                    "gestational_age" ~ "oneway.test",
                    "birth_weight" ~ "oneway.test")) |> 
  modify_fmt_fun(
    p.value = function(x) {
      ifelse(
        x < 0.0001, 
        "<0.0001", 
        format(round(x, 4), nsmall = 4))  # Show 4 decimal places
    }
  )


```


## Associated groups and factors: multinomial regression

### Integrate GBTM uncertainty

We cannot classify the participants with 100% certainty into the various trajectories. We take that into account by using the posterior probabilities as weights in the multinomial logistic regression.

```{r}

long_data <- Leicester_data_weighted |>
  filter(is.na(class)==FALSE) |> 
  pivot_longer(
    cols = starts_with("prob"),
    names_to = "latent_class",
    values_to = "uncertainty_weight"
  )

long_data <- long_data |> 
  mutate(latent_class = factor(latent_class, levels = c("prob4", "prob2", "prob1",  "prob5",  "prob3"), labels = c("Stable normal BMI", "Persistent low BMI", "Early overweight resolving",  "Childhood onset obesity", "Adolescent onset overweight")),
         latent_class = relevel(latent_class, ref = "Stable normal BMI"))

long_data <- long_data |> 
  mutate(total_weight = ipweight_stabilized*uncertainty_weight)

```

### eTable 8: Univariable, unadjusted

#### Sociodemographic

sex

```{r}

model <- multinom(latent_class ~ sex, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_sex <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_sex

```



Birth year

```{r}
#| warning: false

long_data <- long_data |> 
  mutate(birth_year_cat = factor(birth_year_cat,
                                 levels = c("1993-1997", "1985-1989")))

model <- multinom(latent_class ~ birth_year_cat, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_birth_year <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_birth_year

```

Ethnicity

```{r}
#| warning: false

model <- multinom(latent_class ~ ethnicity_cohort, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary



table_summary_ethnicity <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_ethnicity

```

Parental education

```{r}
#| warning: false

long_data <- long_data |> 
  mutate(parental_education_cat = case_when(parental_education_cat=="unkown" ~ NA,
                                            TRUE ~ parental_education_cat))


model <- multinom(latent_class ~ parental_education_cat, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_parental_education <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_parental_education

```


Urbanization

```{r}

model <- multinom(latent_class ~ urban_rural, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_urban_rural <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_urban_rural


```


Townsend 

```{r}
#| warning: false

long_data <- long_data |> 
  mutate(townsend_2001_quintile = factor(townsend_2001_quintile, levels = c(1,2,3,4,5)))


model_t1 <- multinom(latent_class ~ townsend_2001_quintile, data = long_data, weights = total_weight)
model_summary <- tidy(model_t1, conf.int = TRUE, exponentiate = TRUE)
model_summary

model_t2 <- multinom(latent_class ~ townsend_2001, data = long_data, weights = total_weight)
model_summary <- tidy(model_t2, conf.int = TRUE, exponentiate = TRUE)
model_summary

model_t3 <- multinom(latent_class ~ townsend_2001 + I(townsend_2001^2), data = long_data, weights = total_weight)
model_summary <- tidy(model_t3, conf.int = TRUE, exponentiate = TRUE)
model_summary

AIC(model_t1, model_t2, model_t3)
BIC(model_t1, model_t2, model_t3)

model_t1 <- multinom(latent_class ~ townsend_2001_quintile, data = long_data, weights = total_weight)
model_summary <- tidy(model_t1, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_townsend <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_townsend






```




#### Perinatal

Birth order

```{r}



model <- multinom(latent_class ~ birth_order_cat, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_birth_order <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_birth_order

```



Age mother birth

```{r}
#| warning: false

Leicester_data_weighted |>
  filter(!is.na(class)) |>
  summarize(
    sd_val = sqrt(wtd.var(age_mother_delivery, weights = ipweight_stabilized, na.rm = TRUE))
  ) |>
  pull(sd_val)

model <- multinom(latent_class ~ scale(age_mother_delivery), data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_age_mother <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_age_mother

```


Gestational age

```{r}
#| warning: false

Leicester_data_weighted |>
  filter(!is.na(class)) |>
  summarize(
    sd_val = sqrt(wtd.var(gestational_age, weights = ipweight_stabilized, na.rm = TRUE))
  ) |>
  pull(sd_val)

model <- multinom(latent_class ~ scale(gestational_age), data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_gestational_age <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_gestational_age

```

Birth_weight

```{r}
#| warning: false

Leicester_data_weighted |>
  filter(!is.na(class)) |>
  summarize(
    sd_val = sqrt(wtd.var(birth_weight, weights = ipweight_stabilized, na.rm = TRUE))
  ) |>
  pull(sd_val)

model <- multinom(latent_class ~ scale(birth_weight), data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_birth_weight <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_birth_weight

```

#### Lifestyle 

Smoking pregnancy

```{r}
#| warning: false



model <- multinom(latent_class ~ smoking_mother_pregnancy, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_smoking_pregnancy <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_smoking_pregnancy

```

Breast feeding

```{r}
#| warning: false

model <- multinom(latent_class ~ breast_feeding, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_breast_feeding <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_breast_feeding

```





Second hand smoke

```{r}
#| warning: false


model <- multinom(latent_class ~ smoking_others_ever, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_smoking_others_ever <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_smoking_others_ever

```


Vigorous physical activity

```{r}
#| warning: false

model <- multinom(latent_class ~ physical_activity_cat_minimal, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_physical_activity <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_summary_physical_activity

```

#### Diet composition

Fresh fruit

```{r}
#| warning: false


long_data <- long_data |> 
  mutate(fresh_fruit = factor(fresh_fruit_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))


model <- multinom(latent_class ~ fresh_fruit, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_fresh_fruit <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  ) |> filter(term == "fresh_fruitless than once daily")

table_summary_fresh_fruit


model_summary_fresh_fruit <- model_summary |> 
  filter(term !="(Intercept)") |> 
  filter(term == "fresh_fruitless than once daily") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)

```

Vegetables

```{r}
#| warning: false


long_data <- long_data |> 
  mutate(vegetables = factor(vegetables_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))


model <- multinom(latent_class ~ vegetables, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_vegetables <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  ) |> filter(term == "vegetablesless than once daily")

table_summary_vegetables


model_summary_vegetables <- model_summary |> 
  filter(term !="(Intercept)") |> 
  filter(term == "vegetablesless than once daily") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)

```



Crisps

```{r}
#| warning: false

long_data <- long_data |> 
  mutate(crisps = factor(crisps_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))

model <- multinom(latent_class ~ crisps, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_crisps <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  ) |> filter(term == "crispsless than once daily")

table_summary_crisps


model_summary_crisps <- model_summary |> 
  filter(term !="(Intercept)") |> 
  filter(term == "crispsless than once daily") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)

```



Chocolates

```{r}
#| warning: false


long_data <- long_data |> 
  mutate(chocolates = factor(chocolate_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))



model <- multinom(latent_class ~ chocolates, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_summary_chocolates <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  ) |> filter(term == "chocolatesless than once daily")

table_summary_chocolates


model_summary_chocolates <- model_summary |> 
  filter(term !="(Intercept)") |> 
  filter(term == "chocolatesless than once daily") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)

```







#### Combined table

```{r}

table_multinomial_regression_univariable <- table_summary_sex |> 
    bind_rows(table_summary_birth_year, table_summary_ethnicity, table_summary_parental_education, table_summary_urban_rural, table_summary_townsend, 
              table_summary_birth_order, table_summary_age_mother, table_summary_smoking_pregnancy, table_summary_gestational_age, table_summary_birth_weight, table_summary_breast_feeding,
              table_summary_smoking_others_ever, table_summary_physical_activity, table_summary_fresh_fruit, table_summary_vegetables, table_summary_crisps, table_summary_chocolates)

```


### eTable 9: At-risk groups: adjusted sociodemographics and perinatal


```{r}

cat("number of participants:", Leicester_data |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(ethnicity_cohort) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight)) |> nrow())

Leicester_data |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(ethnicity_cohort) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight)) |> select(class) |> tbl_summary()

model <- multinom(latent_class ~ sex + birth_year_cat + ethnicity_cohort + parental_education_cat + urban_rural + townsend_2001_quintile + birth_order_cat + I(scale(age_mother_delivery)) + I(scale(gestational_age)) + I(scale(birth_weight)), data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_multinomial_regression_at_risk <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_multinomial_regression_at_risk


model_summary_at_risk <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)


```


### Figure 3: At-risk groups

Prepare data for forrest plot

```{r}

# combine model summaries
table_plot_multinomial_regression_at_risk <- model_summary_at_risk


```



```{r}

table_plot_multinomial_regression_at_risk <- table_plot_multinomial_regression_at_risk |> 
  mutate(term = case_when(term == "sexfemale" ~ "Female",
                          term == "birth_year_cat1985-1989" ~ "'1985–1989'",
                          term == "ethnicity_cohortSouth Asian" ~ "'South Asian'",
                          term == "urban_ruralrural" ~ "'Suburbs and rural'",
                          term == "townsend_2001_quintile2" ~ "'Quintile 2'",
                          term == "townsend_2001_quintile3" ~ "'Quintile 3'",
                          term == "townsend_2001_quintile4" ~ "'Quintile 4'",
                          term == "townsend_2001_quintile5" ~ "'Quintile 5'",
                          term == "parental_education_catfurther education" ~ "Further",
                          term == "parental_education_cathigher education" ~ "Higher",
                          term == "I(scale(age_mother_delivery))" ~ "'Per 1 SD increase: 5.3 years'",
                          term == "I(scale(gestational_age))" ~ "'Per 1 SD increase: 1.9 weeks'",
                          term == "I(scale(birth_weight))" ~ "'Per 1 SD increase: 566 grams'",
                          term == "birth_order_catsecond child" ~ "'Second child'",
                          term == "birth_order_catthird or later child" ~ "'Third or later child'"))

table_regression_reference <- data.frame(class = c("variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable"),
                                         term = c("bold(Sex)*', ref: Male'",
                                                  "bold('Birth year')*', ref: 1993–1997'",
                                                  "bold(Ethnicity)*', ref: White'",
                                                  "bold('Degree of urbanization')*', ref: Inner city'",
                                                  "bold('Townsend index')*', ref: Quintile 1'",
                                                  "bold('Parental education')*', ref: Compulsory'",
                                                  
                                                  "bold('Birth order')*', ref: First child'",
                                                  "bold('Age of mother at birth')",
                                                  "bold('Gestational age')",
                                                  "bold('Birth weight')"),
                                         aOR = 1,
                                         lower_ci = NA,
                                         upper_ci = NA)

table_plot_multinomial_regression_at_risk <- table_plot_multinomial_regression_at_risk |> 
  bind_rows(table_regression_reference) |> 
mutate(term = factor(term,
                     levels = c(
                       "'Per 1 SD increase: 566 grams'",
                       "bold('Birth weight')",
                       "'Per 1 SD increase: 1.9 weeks'",
                       "bold('Gestational age')",
                       "'Per 1 SD increase: 5.3 years'",
                       "bold('Age of mother at birth')",
                       "'Third or later child'",
                       "'Second child'",
                       "bold('Birth order')*', ref: First child'",
                       
                       "'Quintile 5'",
                       "'Quintile 4'",
                       "'Quintile 3'",
                       "'Quintile 2'",
                       "bold('Townsend index')*', ref: Quintile 1'",
                       "'Suburbs and rural'",
                       "bold('Degree of urbanization')*', ref: Inner city'",
                       "Higher",
                       "Further",
                       "bold('Parental education')*', ref: Compulsory'",
                       "'South Asian'",
                       "bold(Ethnicity)*', ref: White'",
                       "'1985–1989'",
                       "bold('Birth year')*', ref: 1993–1997'",
                       "Female",
                       "bold(Sex)*', ref: Male'"
                     ))) |> 
  mutate(group = case_when(term == "bold(Sex)*', ref: Male'" ~ "Sociodemographic",
                           term == "Female"~ "Sociodemographic",
                           term == "bold('Birth year')*', ref: 1993–1997'" ~ "Sociodemographic",
                           term == "'1985–1989'" ~ "Sociodemographic",
                           term == "bold(Ethnicity)*', ref: White'"~ "Sociodemographic",
                           term == "'South Asian'"~ "Sociodemographic",
                           term == "bold('Degree of urbanization')*', ref: Inner city'"~ "Sociodemographic",
                           term == "'Suburbs and rural'"~ "Sociodemographic",
                           term == "bold('Townsend index')*', ref: Quintile 1'"~ "Sociodemographic",
                           term == "'Quintile 2'"~ "Sociodemographic",
                           term == "'Quintile 3'"~ "Sociodemographic",
                           term == "'Quintile 4'"~ "Sociodemographic",
                           term == "'Quintile 5'"~ "Sociodemographic",
                           term == "bold('Parental education')*', ref: Compulsory'"~ "Sociodemographic",
                           term == "Further"~ "Sociodemographic",
                           term == "Higher"~ "Sociodemographic",
                           term == "bold('Age of mother at birth')"~ "Birth related",
                           term == "'Per 1 SD increase: 5.3 years'"~ "Birth related",
                           term == "bold('Gestational age')"~ "Birth related",
                           term == "'Per 1 SD increase: 1.9 weeks'"~ "Birth related",
                           term == "bold('Birth weight')" ~ "Birth related",
                           term == "'Per 1 SD increase: 566 grams'"~ "Birth related",
                           term == "bold('Birth order')*', ref: First child'" ~ "Birth related",
                           term == "'Second child'"~ "Birth related",
                           term == "'Third or later child'"~ "Birth related")) |> 
  mutate(hide = factor(case_when(class == "reference" | class == "variable" ~ "Yes",
                          TRUE ~ "No"), levels = c("Yes", "No")))
  

table_plot_multinomial_regression_at_risk <- table_plot_multinomial_regression_at_risk |> 
  mutate(transparency = case_when(term == "bold('Age of mother at birth')" ~ "transparent",
                                  term == "bold('Gestational age')" ~ "transparent",
                                  term == "bold('Birth weight')" ~ "transparent",
                                  TRUE ~ "normal"))

```

#### Plot by trajectory

```{r}

plot_forrest_multinomial_below_normal <- table_plot_multinomial_regression_at_risk |>
  filter(class == "Persistent low BMI" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency), size = 1, position = position_dodge(width = 1), colour = unibeIce()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeIce()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_below_normal


plot_forrest_multinomial_early_overweight <- table_plot_multinomial_regression_at_risk |>
  filter(class == "Early overweight resolving" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeMustard()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeMustard()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_early_overweight


plot_forrest_multinomial_school_obesity <- table_plot_multinomial_regression_at_risk |>
  filter(class == "Childhood onset obesity" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeRed()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeRed()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_school_obesity

plot_forrest_multinomial_adolescent_overweight <- table_plot_multinomial_regression_at_risk |>
  filter(class == "Adolescent onset overweight" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeApricot()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeApricot()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_adolescent_overweight



y_axis_plot <- table_plot_multinomial_regression_at_risk |>
  filter(class == "Childhood onset obesity" | class == "variable") |> 
  ggplot(aes(x = 1, y = term)) +
  geom_point(size = 1, position = position_dodge(width = 1), alpha = 0) +
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.title = element_blank(),
        axis.text.y = element_text(size = 8, color = "black"),
        axis.text.x = element_blank(),
        legend.position = "none",
        plot.margin = margin(l = 0.25, b = 0, t = 0, r = 0, unit = "cm"))

y_axis_plot


plot_forrest_multinomial_at_risk <- plot_grid(y_axis_plot,
                                      plot_forrest_multinomial_below_normal + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_early_overweight + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_school_obesity + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_adolescent_overweight + theme(axis.text.y = element_blank()),
                                      ncol = 5,
                                      align = "h",
                                      labels = c(" ", "Persistent low BMI", "Early overweight resolving", "Childhood onset obesity", "Adolescent onset overweight"),
                                      label_size = 10,
                                      label_x = 0.5,
                                      hjust = 0.5)

ggsave(plot_forrest_multinomial_at_risk, filename = here("output", "figures", "plot_forrest_multinomial_at_risk.svg"), device = "svg", units = "cm", bg = "white", width = 28, height = 11.5)


```


### eTable 10: Risk factors: fully adjusted


```{r}

long_data <- long_data |> 
  mutate(breast_feeding = factor(breast_feeding,
                                             levels = c("yes", "no")))

long_data <- long_data |> 
  mutate(fresh_fruit = factor(fresh_fruit_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))

cat("number of participants:", Leicester_data |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(ethnicity_cohort) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(smoking_mother_pregnancy) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> nrow())

Leicester_data |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(ethnicity_cohort) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(smoking_mother_pregnancy) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> select(class) |> tbl_summary()

model <- multinom(latent_class ~ sex + birth_year_cat + ethnicity_cohort + parental_education_cat + urban_rural + townsend_2001_quintile + birth_order_cat + I(scale(age_mother_delivery)) + I(scale(gestational_age)) + I(scale(birth_weight)) + smoking_mother_pregnancy + breast_feeding + smoking_others_ever + physical_activity_cat_minimal + fresh_fruit, data = long_data, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_multinomial_regression_risk_factors <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2), "–", format(round2(`conf.high`, 2), nsmall=2), "]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_multinomial_regression_risk_factors


model_summary_risk_factors <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)


```

### Figure 4: Risk factors

Prepare data for forrest plot

```{r}

# combine model summaries
table_plot_multinomial_regression_risk_factors <- model_summary_risk_factors


```



```{r}

table_plot_multinomial_regression_risk_factors <- table_plot_multinomial_regression_risk_factors |> 
  mutate(term = case_when(term == "sexfemale" ~ "Female",
                          term == "birth_year_cat1985-1989" ~ "'1985–1989'",
                          term == "ethnicity_cohortSouth Asian" ~ "'South Asian'",
                          term == "urban_ruralrural" ~ "'Suburbs and rural'",
                          term == "townsend_2001_quintile2" ~ "'Quintile 2'",
                          term == "townsend_2001_quintile3" ~ "'Quintile 3'",
                          term == "townsend_2001_quintile4" ~ "'Quintile 4'",
                          term == "townsend_2001_quintile5" ~ "'Quintile 5'",
                          term == "parental_education_catfurther education" ~ "Further",
                          term == "parental_education_cathigher education" ~ "Higher",
                          term == "I(scale(age_mother_delivery))" ~ "'Per 1 SD increase: 5.3 years'",
                          term == "I(scale(gestational_age))" ~ "'Per 1 SD increase: 1.9 weeks'",
                          term == "I(scale(birth_weight))" ~ "'Per 1 SD increase: 566 grams'",
                          term == "birth_order_catsecond child" ~ "'Second child'",
                          term == "birth_order_catthird or later child" ~ "'Third or later child'",
                          term == "smoking_mother_pregnancyyes" ~ "Yes",
                          term == "breast_feedingno" ~ "No",
                          term == "physical_activity_cat_minimal>= 4 hrs per week" ~ "'≥4 hrs/week'",
                          term == "fresh_fruitless than once daily" ~ "'Less than once daily'",
                          term == "smoking_others_everyes" ~ "' Yes'"))

table_regression_reference <- data.frame(class = c("variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable",
                                                   "variable"),
                                         term = c("bold(Sex)*', ref: Male'",
                                                  "bold('Birth year')*', ref: 1993–1997'",
                                                  "bold(Ethnicity)*', ref: White'",
                                                  "bold('Degree of urbanization')*', ref: Inner city'",
                                                  "bold('Townsend index')*', ref: Quintile 1'",
                                                  "bold('Parental education')*', ref: Compulsory'",
                                                  
                                                  "bold('Birth order')*', ref: First child'",
                                                  "bold('Age of mother at birth')",
                                                  "bold('Gestational age')",
                                                  "bold('Birth weight')",
                                                  
                                                  "bold('Smoking during pregnancy')*', ref: No'",
                                                  "bold(Breastfeeding)*', ref: Yes'",
                                                  "bold('Secondhand tobacco smoke')*', ref: No'",
                                                  "bold('Physical activity')*', ref: 0–3 hrs/week'",
                                                  "bold('Fresh fruit intake')*', ref: Once daily or more'"),
                                         aOR = 1,
                                         lower_ci = NA,
                                         upper_ci = NA)

table_plot_multinomial_regression_risk_factors <- table_plot_multinomial_regression_risk_factors |> 
  bind_rows(table_regression_reference) |> 
mutate(term = factor(term,
                     levels = c(
                       "'Less than once daily'",
                       "bold('Fresh fruit intake')*', ref: Once daily or more'",
                       "'≥4 hrs/week'",
                       "bold('Physical activity')*', ref: 0–3 hrs/week'",
                       "' Yes'",
                       "bold('Secondhand tobacco smoke')*', ref: No'",
                       "No",
                       "bold(Breastfeeding)*', ref: Yes'",
                       "Yes",
                       "bold('Smoking during pregnancy')*', ref: No'",
                       
                       "'Per 1 SD increase: 566 grams'",
                       "bold('Birth weight')",
                       "'Per 1 SD increase: 1.9 weeks'",
                       "bold('Gestational age')",
                       "'Per 1 SD increase: 5.3 years'",
                       "bold('Age of mother at birth')",
                       "'Third or later child'",
                       "'Second child'",
                       "bold('Birth order')*', ref: First child'",
                       
                       "'Quintile 5'",
                       "'Quintile 4'",
                       "'Quintile 3'",
                       "'Quintile 2'",
                       "bold('Townsend index')*', ref: Quintile 1'",
                       "'Suburbs and rural'",
                       "bold('Degree of urbanization')*', ref: Inner city'",
                       "Higher",
                       "Further",
                       "bold('Parental education')*', ref: Compulsory'",
                       "'South Asian'",
                       "bold(Ethnicity)*', ref: White'",
                       "'1985–1989'",
                       "bold('Birth year')*', ref: 1993–1997'",
                       "Female",
                       "bold(Sex)*', ref: Male'"
                     ))) |> 
  mutate(group = case_when(term == "bold(Sex)*', ref: Male'" ~ "Sociodemographic",
                           term == "Female"~ "Sociodemographic",
                           term == "bold('Birth year')*', ref: 1993–1997'" ~ "Sociodemographic",
                           term == "'1985–1989'" ~ "Sociodemographic",
                           term == "bold(Ethnicity)*', ref: White'"~ "Sociodemographic",
                           term == "'South Asian'"~ "Sociodemographic",
                           term == "bold('Degree of urbanization')*', ref: Inner city'"~ "Sociodemographic",
                           term == "'Suburbs and rural'"~ "Sociodemographic",
                           term == "bold('Townsend index')*', ref: Quintile 1'"~ "Sociodemographic",
                           term == "'Quintile 2'"~ "Sociodemographic",
                           term == "'Quintile 3'"~ "Sociodemographic",
                           term == "'Quintile 4'"~ "Sociodemographic",
                           term == "'Quintile 5'"~ "Sociodemographic",
                           term == "bold('Parental education')*', ref: Compulsory'"~ "Sociodemographic",
                           term == "Further"~ "Sociodemographic",
                           term == "Higher"~ "Sociodemographic",
                           term == "bold('Smoking during pregnancy')*', ref: No'"~ "Birth related",
                           term == "Yes"~ "Birth related",
                           term == "bold('Age of mother at birth')"~ "Birth related",
                           term == "'Per 1 SD increase: 5.3 years'"~ "Birth related",
                           term == "bold('Gestational age')"~ "Birth related",
                           term == "'Per 1 SD increase: 1.9 weeks'"~ "Birth related",
                           term == "bold('Birth weight')" ~ "Birth related",
                           term == "'Per 1 SD increase: 566 grams'"~ "Birth related",
                           term == "bold('Birth order')*', ref: First child'" ~ "Birth related",
                           term == "'Second child'"~ "Birth related",
                           term == "'Third or later child'"~ "Birth related",
                           term == "bold(Breastfeeding)*', ref: Yes'"~ "Birth related",
                           term == "No"~ "Birth related",
                           term == "bold('Secondhand tobacco smoke')*', ref: No'" ~ "Lifestyle",
                           term == "' Yes'" ~ "Lifestyle",
                           term == "bold('Physical activity')*', ref: 0–3 hrs/week'" ~ "Lifestyle",
                           term == "'≥4 hrs/week'" ~ "Lifestyle",
                           term == "bold('Fresh fruit intake')*', ref: Once daily or more'" ~ "Lifestyle",
                           term == "'Less than once daily'" ~ "Lifestyle")) |> 
  mutate(hide = factor(case_when(class == "reference" | class == "variable" ~ "Yes",
                          TRUE ~ "No"), levels = c("Yes", "No")))
  

table_plot_multinomial_regression_risk_factors <- table_plot_multinomial_regression_risk_factors |> 
  mutate(transparency = case_when(term == "bold('Age of mother at birth')" ~ "transparent",
                                  term == "bold('Gestational age')" ~ "transparent",
                                  term == "bold('Birth weight')" ~ "transparent",
                                  TRUE ~ "normal"))

```

#### Plot by trajectory

```{r}

plot_forrest_multinomial_below_normal <- table_plot_multinomial_regression_risk_factors |>
  filter(class == "Persistent low BMI" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency), size = 1, position = position_dodge(width = 1), colour = unibeIce()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeIce()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_below_normal


plot_forrest_multinomial_early_overweight <- table_plot_multinomial_regression_risk_factors |>
  filter(class == "Early overweight resolving" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeMustard()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeMustard()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_early_overweight


plot_forrest_multinomial_school_obesity <- table_plot_multinomial_regression_risk_factors |>
  filter(class == "Childhood onset obesity" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeRed()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeRed()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_school_obesity

plot_forrest_multinomial_adolescent_overweight <- table_plot_multinomial_regression_risk_factors |>
  filter(class == "Adolescent onset overweight" | class == "variable") |> 
  ggplot(aes(x = aOR, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0.25, color = "black") +
  geom_vline(xintercept = 4, color = "black") +
  geom_point(aes(alpha = transparency),size = 1, position = position_dodge(width = 1), colour = unibeApricot()) +
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci),
                 height = 0.25, linewidth = 0.25,
                 position = position_dodge(width = 1),
                 colour = unibeApricot()) +
  scale_x_continuous(breaks = c(0.25, 0.5, 1, 2, 4), transform = "log", expand = c(0,0)) +
  coord_cartesian(xlim = c(0.25, 4), expand = FALSE, clip = "on")+
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  scale_alpha_manual(breaks = c("transparent", "normal"), values = c(0, 100), guide = "none")+
  labs(x = "**aOR** (log scale)", y = "") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_markdown(size = 8),
        axis.text.y = element_text(size = 8, color = "black", margin = margin(0, 0, 0, 0)),
        axis.text.x = element_text(size = 8, color = "black"),
        legend.position = "none",
        plot.margin = margin(t = 0.75, r = 0.7, l = 0.4, b = 0.5, unit = "cm"))
    


plot_forrest_multinomial_adolescent_overweight



y_axis_plot <- table_plot_multinomial_regression_risk_factors |>
  filter(class == "Childhood onset obesity" | class == "variable") |> 
  ggplot(aes(x = 1, y = term)) +
  geom_point(size = 1, position = position_dodge(width = 1), alpha = 0) +
  scale_y_discrete(labels = function(x) parse(text = x), expand = c(0,0))+
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        axis.title = element_blank(),
        axis.text.y = element_text(size = 8, color = "black"),
        axis.text.x = element_blank(),
        legend.position = "none",
        plot.margin = margin(l = 0.25, b = 0, t = 0, r = 0, unit = "cm"))

y_axis_plot


plot_forrest_multinomial_risk_factors <- plot_grid(y_axis_plot,
                                      plot_forrest_multinomial_below_normal + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_early_overweight + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_school_obesity + theme(axis.text.y = element_blank()),
                                      plot_forrest_multinomial_adolescent_overweight + theme(axis.text.y = element_blank()),
                                      ncol = 5,
                                      align = "h",
                                      labels = c(" ", "Persistent low BMI", "Early overweight resolving", "Childhood onset obesity", "Adolescent onset overweight"),
                                      label_size = 10,
                                      label_x = 0.5,
                                      hjust = 0.5)

ggsave(plot_forrest_multinomial_risk_factors, filename = here("output", "figures", "plot_forrest_multinomial_risk_factors.svg"), device = "svg", units = "cm", bg = "white", width = 28, height = 17.5)


```




### eTable 11 & 12: Stratified by ethnicity


#### South Asians

```{r}

long_data_asians <- long_data |> filter(ethnicity_cohort == "South Asian")

Leicester_data |> filter(ethnicity_cohort == "South Asian") |> select(class) |> tbl_summary()


```


```{r}

long_data_asians <- long_data_asians |> 
  mutate(parental_education_cat = factor(parental_education_cat,
                                             levels = c("higher education", "further education", "compulsory education")))

long_data_asians <- long_data_asians |> 
  mutate(townsend_2001_quintile = factor(townsend_2001_quintile,
                                             levels = c("5", "4", "3", "2", "1")))


long_data_asians <- long_data_asians |> 
  mutate(fresh_fruit = factor(fresh_fruit_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))

cat("number of participants:", Leicester_data |> filter(ethnicity_cohort == "South Asian") |> filter(!is.na(class) & !is.na(parental_education_cat) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> nrow())

Leicester_data |> filter(ethnicity_cohort == "South Asian") |> filter(!is.na(class) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> select(class) |> tbl_summary()

model <- multinom(latent_class ~ sex + parental_education_cat + townsend_2001_quintile + birth_order_cat + I(scale(age_mother_delivery)) + I(scale(gestational_age)) + I(scale(birth_weight)) + breast_feeding + smoking_others_ever + physical_activity_cat_minimal + fresh_fruit, data = long_data_asians, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_multinomial_regression_risk_factors_SA <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2),"–",format(round2(`conf.high`,2),nsmall=2),"]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_multinomial_regression_risk_factors_SA


model_summary_risk_factors_SA <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)


```
#### Whites

```{r}

long_data_whites <- long_data |> filter(ethnicity_cohort == "White")

Leicester_data |> filter(ethnicity_cohort == "White") |> select(class) |> tbl_summary()


```


```{r}

long_data_whites <- long_data_whites |> 
  mutate(fresh_fruit = factor(fresh_fruit_cat_mean,
                              levels = c("Once daily or more", "less than once daily")))



cat("number of participants:", Leicester_data |> filter(ethnicity_cohort == "White") |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(smoking_mother_pregnancy) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> nrow())

Leicester_data |> filter(ethnicity_cohort == "White") |> filter(!is.na(class) & !is.na(birth_year_cat) & !is.na(parental_education_cat) & !is.na(urban_rural) & !is.na(townsend_2001_quintile) & !is.na(birth_order_cat) & !is.na(age_mother_delivery) & !is.na(gestational_age) & !is.na(birth_weight) & !is.na(smoking_mother_pregnancy) & !is.na(breast_feeding) & !is.na(smoking_others_ever) & !is.na(physical_activity_cat_minimal) & !is.na(fresh_fruit_cat)) |> select(class) |> tbl_summary()

model <- multinom(latent_class ~ sex + birth_year_cat + parental_education_cat + urban_rural + townsend_2001_quintile + birth_order_cat + I(scale(age_mother_delivery)) + I(scale(gestational_age)) + I(scale(birth_weight)) + smoking_mother_pregnancy + breast_feeding + smoking_others_ever + physical_activity_cat_minimal + fresh_fruit, data = long_data_whites, weights = total_weight)
model_summary <- tidy(model, conf.int = TRUE, exponentiate = TRUE)
model_summary

table_multinomial_regression_risk_factors_WH <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(`OR [95% CI]` = paste0(round2(estimate, 2), " [", format(round2(`conf.low`, 2), nsmall=2),"–",format(round2(`conf.high`,2),nsmall=2),"]")) |> 
  select(`y.level`, `term`, `OR [95% CI]`) |> 
  pivot_wider(
    names_from = y.level,
    values_from = `OR [95% CI]` 
  )

table_multinomial_regression_risk_factors_WH


model_summary_risk_factors_WH <- model_summary |> 
  filter(term !="(Intercept)") |> 
  mutate(aOR = estimate,
         lower_ci = conf.low,
         upper_ci = conf.high,
         class = y.level) |> 
  select(class, term, aOR, lower_ci, upper_ci)


```