---
title: "1 class models with raw bmi and spline patterns"
author: "Lorenz Leuenberger"
date: "`r date()`"
title-block-banner: "#e4003c"
format: 
  html:
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    df-print: paged
    embed-resources: true
---

# Introduction

To model the higher class models (n >= 2) on Ubelix we first need to specify and model the baseline models (n = 1). Because they are not compute intensive we can run them locally.

The modelling approach and the specification and naming of the models is described in the Concept sheet.

We will first model sex and ethnicity separately and then combine.
We will model raw bmi.

# Import data

libraries

```{r}

library(tidyverse)
library(here)
library(lcmm)

```


```{r}

data <- read.csv(file = here("data", "processed", "Leicester_anthro_pseudonymized_18.csv"))

```


# Sex

Model female and male data separately

## Female

Select data

```{r}

data_female <- data %>% 
  filter(sex == "f")

```

### Estimate models

```{r}

bmi_m1r_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

bmi_m1s_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

bmi_m1t_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

bmi_m1u_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

bmi_m1v_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

bmi_m1w_female <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_female)

```


### Save models

Save models to .Rdata file

```{r}

save(data_female, bmi_m1r_female, bmi_m1s_female, bmi_m1t_female, bmi_m1u_female, bmi_m1v_female, bmi_m1w_female, file = here("ubelix", "input", "final_splines", "bmi_data_and_1-class_female.Rdata"))

```

## Male


Select data

```{r}

data_male <- data %>% 
  filter(sex == "m")

```

### Estimate models

```{r}

bmi_m1r_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

bmi_m1s_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

bmi_m1t_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

bmi_m1u_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

bmi_m1v_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

bmi_m1w_male <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_male)

```


### Save models

Save models to .Rdata file

```{r}

save(data_male, bmi_m1r_male, bmi_m1s_male, bmi_m1t_male, bmi_m1u_male, bmi_m1v_male, bmi_m1w_male, file = here("ubelix", "input", "final_splines", "bmi_data_and_1-class_male.Rdata"))

```

# Ethnicity

Model European and Non-European ethnicity data separately

## Europe

Select data

```{r}

data_europe <- data %>% 
  filter(ethnicity_cohort == "whites")

```

### Estimate models


```{r}

bmi_m1r_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

bmi_m1s_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

bmi_m1t_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

bmi_m1u_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

bmi_m1v_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

bmi_m1w_europe <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_europe)

```


### Save models

Save models to .Rdata file

```{r}

save(data_europe, bmi_m1r_europe, bmi_m1s_europe, bmi_m1t_europe, bmi_m1u_europe, bmi_m1v_europe, bmi_m1w_europe, file = here("ubelix", "input", "final_splines", "bmi_data_and_1-class_europe.Rdata"))

```

## Asia

Select data

```{r}

data_asia <- data %>% 
  filter(ethnicity_cohort == "asians")

```

### Estimate models


```{r}

bmi_m1r_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

bmi_m1s_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

bmi_m1t_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

bmi_m1u_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

bmi_m1v_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

bmi_m1w_asia <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_asia)

```


### Save models

Save models to .Rdata file

```{r}

save(data_asia, bmi_m1r_asia, bmi_m1s_asia, bmi_m1t_asia, bmi_m1u_asia, bmi_m1v_asia, bmi_m1w_asia, file = here("ubelix", "input", "final_splines", "bmi_data_and_1-class_asia.Rdata"))

```

# Total

Model all data together


Select data

```{r}

data_total <- data

```

### Estimate models


```{r}

bmi_m1r_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

bmi_m1s_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

bmi_m1t_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 9.80449), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

bmi_m1u_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ -1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

bmi_m1v_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

bmi_m1w_total <- hlme(fixed = bmi ~ splines::bs(age_years, knots = c(0.6650824, 10.34597, 13.7964), degree = 2),
                       random = ~ 1 + age_years,
                       subject = "participant_id",
                       ng = 1,
                       nwg = FALSE,
                       idiag = FALSE,
                       data = data_total)

```


### Save models

Save models to .Rdata file

```{r}

save(data_total, bmi_m1r_total, bmi_m1s_total, bmi_m1t_total, bmi_m1u_total, bmi_m1v_total, bmi_m1w_total, file = here("ubelix", "input", "final_splines", "bmi_data_and_1-class_total.Rdata"))

```


